---
title: Estimating Treatment Effects in Longitudinal Clinical Trials with Missing Data 
authors:
  - name: Amy Browne
    thanks: The authors would like to thank Neil O'Leary from the University of Galway for his valuable guidance, feedback, and support throughout the development of this project
    department: Department of Methametics
    affiliation: University of Galway
    email: k.yip2@universityofgalway.ie
  - name: Tsz Mang Yip
    department: Department of Methametics
    affiliation: University of Galway
    email: a.browne47@universityofgalway.ie
abstract:
  - Motivation. Missing data is a pervasive issue in longitudinal clinical trials, risking bias and reduced power. This study compares several statistical methods for estimating treatment effects in the presence of missing data.
  - Result. No meaningful differences between methods for both data sets. To understanding there are many ways to handle missing data, and it is crutial to understand the mechanism to make corret choise under circumstances.
  - Supplement information. Code Available at https://github.com/amydebrun/Missing-Data
keywords:
  - Missing Data
  - Longitudinal Data
  - Randomized Clinical Trial
  - Real World Data Analysis
bibliography: references.bib
biblio-style: unsrt
output: rticles::arxiv_article
header-includes:
  - \newcommand{\pandocbounded}[1]{#1}
---

```{r setup, include=FALSE, cache=TRUE}
knitr::opts_chunk$set(echo = FALSE, warning=FALSE)
library(mice)
library(tidyverse)
library(haven)
library(lme4)
library(broom.mixed)
library(meta)
library(dplyr)
library(naniar)
library(tidymodels)
library(gtsummary)
library(tinytex)
suppressWarnings({
  source("R scripts/Functions.R")
  source("R scripts/VITAL data formats.R")
  source("R scripts/Acupuncture data formats.R")
  source("R scripts/Summary analysis.R")
  source("R scripts/Category time analysis.R")
  source("R scripts/Continuous time analysis.R")
  source("R scripts/Different Imputation Method.R")
  source("R scripts/Sens_analysis_cat.R")
  source("R scripts/Sens_analysis_cont.R")
  source("R scripts/Sens_analysis_combined.R")
  source("R scripts/Data_summary.R")
})
```


# Introduction

## Missing data in longitudinal study

  - Clinical trials often face missing outcome data due to dropout or non-response.
  - longitudinal study designs 
  - missing baseline data (vital)
  - losing data, loss of efficiency, bias
  
  
## Missing mechanism
  
  - Missing mechanism MCAR, MAR, MNAR. (Missing Data Mechanisms)
  
Missing data mechanisms are important to consider when choosing which sort of missing data handling method to use. There are three mechanisms which missing data can follow:

- Missing Completely At Random (MCAR)
- Missing At Random (MAR)
- Missing Not At Random (MNAR)

Although they may appear similar at first glance, handling missing data without considering these mechanisms may result in biased estimates and inaccurate conclusions. 

**Missing Completely At Random**

The formal definition of MCAR data is:

$$\quad P(R = 1 \mid Y, X) = P(R = 1)$$
where $R$ is a missing data indicator (1 = Data is observed, 0 = Data is missing), $Y$ represents the variables in which the data is potentially missing and $X$ represents the observed data. The probability of the data is observed given observed data and missing data is the same as the probability of being observed without the given data. This mechanism is considered the easiest to deal with as it does not bias the result although data is rarely MCAR. This can occur due to system failure and some data is deleted accidentally, or else there is issues with the treatment system and data cannot be recorded.    


**Missing At Random**

$$\quad P(R = 1 \mid Y, X) = P(R = 1 \mid X)$$
The probability of data being observed given the rest of the data is the same as the probability being observed given the observed data. In short, the data's missingness is dependent on the observed data. For example, people with a higher body mass index may be more prone to having missing blood pressure data - this is not relative to the missing data. MAR is a more realistic mechanism than MCAR and requires more intensive handling methods. 


**Missing Not At Random**

$$\quad P(R = 1 \mid Y, X) \ne P(R = 1 \mid X)$$
The probability of data being observed is not dependent on the observed data. This mechanism is the most difficult to deal with as it relates to the unobserved data, so producing valid results is a challenge. Certain participants in a general health study may avoid answering questions truthfully about smoking habits or their diet in order to make themselves more appealing. Sensitivity analysis is an option to determine the treatment effect when assuming different mechanisms. 

  - DAG of missing mechanism (randomised)
 
## Current approach and limitations
 
  - Existing literature reveals inconsistent reporting and handling practices.(Power 2014)
  - Consequences: bias, reduced efficiency, invalid conclusions if not handled properly.
  

## Objective of this project

  - Using statistical principled methods
  - Test how different methods perform in estimating treatment effects using real-world data.

# Data sets

We have sourced two different data sets to perform our analysis on. They both have different sample sizes and different levels of missing data which is an advantage as we can conduct analysis in different settings. 

**Acupuncture Data**

*Source*
  
Vickers et al. published a paper in 2004 on a trial they conducted to determine the effect of acupuncture therapy on chronic headache in primary care. Vickers released the dataset from the study in 2006 and is publicly available  to download in excel format from https://pmc.ncbi.nlm.nih.gov/articles/PMC1489946/.

*Study Design*

The study design for the acupuncture trial is a longitudinal randomised controlled trial with two follow up time points. 401 participants were gathered from general practices in England and Wales who suffered from migraines. 

*Data*

13 variables were recorded during the acupuncture trial. 

```{r}
acu_var_table
```

Variables `pk2` and ``pk5` are the follow-up times at 3 months and 12 months. 

```{r}
acu_summary
```


```{r}
acu_sex_plot
```

```{r}
acu_age_plot
```

```{r}
acu_pk1_plot
```

```{r}
acu_pk2_plot
```

```{r}
acu_pk5_plot
```

*Original Estimand*

The aim of the original study was to estimate the effect of acupuncture therapy on chronic headache in comparison to general care. Therefore, our estimand with the sourced data set is the difference in mean headache score at the final time point of 12 months between the acupuncture group and the control group. This is modelled parsimoniously by

$$\small\text{pk5}_i = \beta_0 + \beta_1 \text{group}_i + \beta_2 \text{pk1}_i + \varepsilon_i$$
where

- $pk5_i$ is the headache severity of individual $i$ at 12 months. 
- $\beta_0$ is our intercept parameter when there is no treatment group and no recorded baseline headache severity. 
- $\beta_1$ is the effect of the treatment group. This is our treatment effect of interest. 
- $\beta_2$ is the effect of the baseline headache severity
- $\varepsilon_i$ is the residual error term for individual $i$

**VITAL data**

  - Source
  - Study design
  - Data overview
    - assuming no interaction in VITAL study
  - Estimands in this project

## Missing analysis

**Acupuncture Data**

```{r}
acu_miss_perc_plot
acu_miss_perc_group
acu_miss_pattern_plot
```

**VITAL data**

```{r}
vital_miss_perc_plot
```
```{r}
vital_miss_perc_plot_fish
```
```{r}
vital_miss_pattern_fish
```

  - Missing proportion
  - Missing pattern
  - which methods to use based on data
  
  
  
  
  
  
  
# Methods

## Data wrangling

  - Have data both in wide and long format
  - We have done multi-imputation in wide format then transform for LME
  - Which account for repeat measurements and will work if collecting time are more different

## Changing imputation methods

  - No imputation (complete case, LME)
  - Single imputation (LOCF, mean observation)
  - Multiple imputation (Different method, predictive value or observed value, number of m)
  - Chain equation for multi-variate missingness

## Changing substantial model

  - multiple linear regression
  - LME
  - Change estimand
  - Change linear regression to polynomial, splie...

## Sensitive analysis

  - Violation of MAR possible
  - shifting method (choosing delta, how to model)

# Result

## Comparing different methods

## Changing estimand
  - almost no impact in acupuncture study due to 2 follow-up

## Changing imputation methods

## Sensitive analysis

## Missing data in clinical research

# Discussion

  - No meaningful difference observed
  - limitation: only 2 follow up for acupuncture
  - limitation: VITAL only have weak therapeutic effect
  - further work: simulate complete data, and compare accuracy between methods

# References























# Back up writing

## Missing Data Mechanisms



## Data 

We will analyse 2 longitudinal randomised controlled trials. Both studies are designed differently with two different sample sizes, which is an advantage to our research as we can look to apply these methods in two different settings. 

## Acupuncture for chronic headache in primary care: large, pragmatic, randomised trial (Vickers et al., 2004)

**Study Design**

Vickers et al. conducted a longitudinal randomised controlled trial to determine the effect of acupuncture therapy on headache against a placebo. 

**Characteristics**

table summary

**Missing Data**


-estimands 

## The Effects of Vitamin D and Marine Omega-3 Fatty Acid Supplementation on Chronic Knee Pain in Older US Adults: Results From a Randomized Trial (MacFarlane et al. 2020)

**Study Design**

**Characteristics**

table summary

**Missing Data**

- estimand 

## Methods

## Complete Case Analysis

- method 
mechanism

## Single Imputation

mean
locf 
mechanism

## Maximum Likelihood

- linear mixed effects model
- conditions 
- only suitable with missing outcome 
mechanism

## Multiple Imputation

- chained equations 
mechanism

## Sensitivity Analysis 

We conducted a sensitivity analysis on both data sets to determine how sensitive the estimated in the treatment effect is when we assume the data is Missing Not At Random. 

**Purpose of Sensitivity Analysis**

As mentioned in our background section, it is important to investigate the missingness mechanism of the data to assess the choice of data handling method. Although we can never really confirm what type of missingness mechanism follows, we can use sensitivity analysis to test how the data forms under different missingness assumptions. 

**Delta Adjustment**

This is a common approach of sensitivity analysis. A fixed $\delta$ is added or removed from the values that were imputed using multiple imputation using chained equations. Doing this, we are estimating the effect that treatment had on the patient if they were observed. Providing a range of deltas allows potential for unobserved patients responses to treatment to be worse or better than the observed patients. 

**Choosing δ**

In our analysis, we chose which range of $\delta$ values to use based on the outcome of the observed values of each data set. A quantitative estimate would involve determining the mean value of the outcome and the mean predicted value of the missing outcome. 

**Acupuncture**


```{r}
#categorical acupuncture
delta_result_cat_plot
#continuous acupuncture placebo
delta_results_cont_acu_placebo_plot 
#categorical acu placebo
delta_result_cat_placebo_plot 
#continuous acupuncture
delta_results_cont_acu_plot


delta_results_cont_acu
delta_results_cont_acu_placebo

acu_heatmap

delta_results_cat <- delta_results_cat %>%
  rename(est_cat = est_cat, delta_cat = delta_acu)

delta_results_cont_acu <- delta_results_cont_acu %>%
  rename(delta_cont = delta_acu)
heatmap_data <- merge(delta_results_cat, delta_results_cont_acu, by = NULL)

ggplot(heatmap_data, aes(x = factor(delta_cont), y = factor(delta_cat), fill = est_cont)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(est_cont, 2)), size = 3) +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  labs(
    title = "Acupuncture Sensitivity Analysis (Effect on pk_score)",
    x = "δ for Continuous Outcome (pk_score)",
    y = "δ for Categorical Outcome (pk5)",
    fill = "Estimate (pk_score)"
  ) +
  theme_minimal() +
  theme(aspect.ratio = 1)

```

**VITAL**
```{r}

delta_results_cont_vitd_plot 
delta_results_cont_fishoil_plot
delta_results_cont_vital_placebo_plot
```


```{r}

delta_results_cont_vitd_plot 


delta_results_cont_fishoil
delta_results_cont_vitd
delta_results_cont_vital_placebo
```


```{r}
vital_MI_LME_result_oil_cont
vital_MI_LME_result_vitd_cont

sum(is.na(vital_wide$pain_yr4) & vital_wide$fishoilactive == 0 & vital_wide$vitdactive == 1)
sum(is.na(vital_wide$pain_yr3) & vital_wide$fishoilactive == 0 & vital_wide$vitdactive == 1)
sum(is.na(vital_wide$pain_yr2) & vital_wide$fishoilactive == 0 & vital_wide$vitdactive == 1)
sum(is.na(vital_wide$pain_yr1) & vital_wide$fishoilactive == 0 & vital_wide$vitdactive == 1)
```
```{r}
delta_results_cat_fishoil_plot
delta_results_cat_vitd_plot 
delta_results_cat_vital_placebo_plot
```


```{r}
delta_combined_cont_acu_plot
delta_combined_cat_acu_plot
delta_combined_cont_cat_plot

delta_combined_cont_acu


```






## Results 

forest plots 
table of estimates 
sensitivity analysis 


