---
title: Estimating Treatment Effects in Longitudinal Clinical Trials with Missing Data 
authors:
  - name: Amy Browne
    thanks: 
    department: School of Mathematical and Statistical Sciences
    affiliation: University of Galway
    email: a.browne47@universityofgalway.ie
  - name: Tsz Mang Yip
    department: School of Mathematical and Statistical Sciences
    affiliation: University of Galway
    email: k.yip2@universityofgalway.ie
abstract: |
    \textbf{Motivation} Missing data is a pervasive issue in longitudinal clinical trials, potentially introducing bias and reducing statistical power. This project tested different missing data handling techniques, using two real-world longitudinal randomized clinical trial (RCT) datasets.
    \textbf{Results} No substantial differences were observed between the methods across both datasets.
    \textbf{Supplementary Information} Code is available at \url{https://github.com/amydebrun/Missing-Data}.
keywords:
  - Missing Data
  - Longitudinal Data
  - Randomized Clinical Trial
  - Real World Data Analysis
  - Sensitivity Analysis
  - Multiple Imputation
  - Missingness Mechanisms
bibliography: references.bib
biblio-style: unsrt
output: rticles::arxiv_article
header-includes:
  - \usepackage{amsmath}
  - \newcommand{\pandocbounded}[1]{#1}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning=FALSE)
library(meta)
library(gtsummary)
library(dagitty)
library(ggplot2)
load("report_plots1.RData")
load("report_plots2.RData")
```

\newpage

# Introduction - amy

## Missing data in longitudinal study

*Longitudinal Study Design*

Longitudinal studies are research designs in which data are collected
from the same participants at multiple time points. They provide
valuable insights into how variables change over time—often over several
years or decades. While this design is powerful for understanding causal
relationships and long-term trends, it presents significant challenges
in retaining participants across all waves of data collection.

Longitudinal trials require consistent data collection methods at
predefined intervals. Researchers must work to minimize participant
dropout or non-response. However, due to the extended duration and
complexity of these studies, missing data is almost inevitable.

Similarly, clinical trials which evaluate the effects of biomedical or
behavioral interventions are also vulnerable to missing data.
Participants may miss follow-ups due to illness or other commitments.
This loss of data can compromise the study’s validity. Ignoring missing
data introduces bias, as we’re disregarding potentially informative
cases. For instance, if participants drop out due to adverse effects
from the treatment, the missing data is not random—this introduces
attrition bias, which can distort estimates of the treatment’s effect.

Removing incomplete cases reduces the sample size, which in turn
decreases statistical power, limiting the ability to detect valid
treatment effects.

In statistical software like R, missing values are represented as NA
among the observed values. In longitudinal data, rows of data belonging
to individuals at a certain time-point can be missing. When fitting
models like linear regression, R automatically excludes any subjects
with missing values in either predictor or outcome variables. While
convenient, this method can lead to selection bias if the missingness is
systematic.

## Project Objective

The objective of this project is to estimate treatment effects using
real-world data, with a primary focus on addressing issues that come
with missing data in clinical trials. We will briefly discuss different
assumptions about missing data mechanisms—such as Missing Completely at
Random, Missing at Random, and Missing Not at Random and investigate
their implications for statistical analysis.

While a range of methods are available to handle missing data, they all
vary in levels of complexity and assumptions. This project aims to
explore these methods, understand the reasons behind their differing
performance, and implement them on real-world longitudinal clinical data
to compare their ability to estimate treatment effect under real-world
conditions.

After initial exploration of the different methods, we also decided to
adjust the estimands through data wrangling our chosen datasets to
compare how the estimate differs when the estimand is either categorical
or continuous and to compare the effects using linear regression or
linear mixed effects models. Additionally, we will conduct sensitivity
analyses to investigate how treatment effect estimates change under
different assumptions about the missing data.

## Missingness Mechanisms

Missing data mechanisms are important to consider when choosing which
sort of missing data handling method to use. There are three mechanisms
which missing data can follow:

-   Missing Completely At Random (MCAR)
-   Missing At Random (MAR)
-   Missing Not At Random (MNAR)

Although they may appear similar at first glance, continuing to handle
missing data without considering these mechanisms may still result in
biased estimates and inaccurate conclusions. Missing data mechanisms
appear to be mentioned first by Donald Rubin in 1976. We can formally
define these mechanisms using the following notation;

-   $R$ represents a missing data indicator which when $R=1$ indicates
    observed data and $R=0$ indicates unobserved
-   $X$ represents data that is always observed
-   $Y$ represents data that is potentially missing.

**Missing Completely At Random**

The formal definition of MCAR data is:

$$\quad P(R = 1 \mid Y, X) = P(R = 1)$$ The probability of the data is
observed given observed data and missing data is the same as the
probability of being observed without the given data. This mechanism is
considered the easiest to deal with as it does not bias the result
although data is rarely MCAR. This can occur due to system failure and
some data is deleted accidentally, or else there is issues with the
treatment system and data cannot be recorded. The DAG below visualises
an example of this mechanism based on a randomised trial with
randomisation variable $L$. Missing indicator $R_Y$, which denotes
whether the outcome is observed ($R_Y = 1$) or missing ($R_Y = 0$), is
independent from the observed baseline variables $X_1$ and $X_2$, which
are pre-randomised variables, and outcome variable $Y$ (which
potentially contains missing data).

```{r echo=FALSE}
MCAR_dag
```

**Missing At Random**

$$\quad P(R = 1 \mid Y, X) = P(R = 1 \mid X)$$ The probability of data
being observed given observed and unobserved data is the same as the
probability being observed given the observed data. In short, the
missing data dependends on the observed data. We can say that the
missing data indicator is conditional on another variable.

MAR is a more realistic mechanism than MCAR and requires more intensive
handling methods. Below shows 2 DAGs that shows a general example of a
randomised treatment $L$ on response variable $Y$, which potentially
contains missing data. $X_1$ and $X_2$ are baseline variables, recorded
pre-randomisation. The first DAG shows the MAR mechanism conditioned on
both baseline variables; missingness indicator $R_Y$ is influenced by
baseline variables $X_1$ and $X_2$. The second DAG is similar, but shows
the data is MAR conditional on randomised treatment $L$ and baseline
variable $X_1$. $X_2$ no longer influences the missingness indicator
$R_Y$, but $R_Y$ is still conditioned on $L$ and $X_1$, showing the MAR
assumption still holds given these variables.

```{r echo=FALSE}
MAR_dag1
MAR_dag2
```

**Missing Not At Random**

$$\quad P(R = 1 \mid Y, X) \ne P(R = 1 \mid X)$$ The probability of data
being observed depends both on the observed data and the missing data.
This mechanism is the most difficult to deal with as it relates to the
unobserved data, so producing valid results is a challenge. Certain
participants in a general health study may avoid answering questions
truthfully about smoking habits or their diet in order to make
themselves more appealing. Sensitivity analysis is an option to
determine the treatment effect when assuming different mechanisms. The 2
following DAGs show this mechanism using the above scenarios. In the
first DAGs structure, the potentially missing outcome variable $Y$
directly influences $R_Y$. The outcome $Y$ is directly influenced by the
randomised treatment $L$ and baseline variables $X_1$ and $X_2$, but the
missingness indicator is influenced by the missing outcome data. The
second DAG shows MNAR conditional on $X_2$ only as it influences the
missingness indicator $R_Y$ along with the missing outcome.

```{r echo=FALSE}
MNAR_dag1
MNAR_dag2
```

Although these considerations are important, the mechanism underlying
missing data is fundamentally untestable. Since the true values of
missing observations are never observed, we cannot definitively
determine which missingness assumption holds in practice. There is
limited hypothesis test available to see whether data is MCAR (*Little's
MCAR test*). It is a likelihood-ratio test that groups missing data
patterns, estimates the expected means under a multivariate normal model
using maximum likelihood, and computes a chi-square statistic to assess
model fit. The decision lies under the hypotheses;

-   $H_0$: Data is MCAR
-   $H_1$: Data is MAR or MNAR

If the p-value resulting from the chi-square distribution is greater
than 0.05, we fail to reject the null hypothesis. It is limited as it
does not necessarily confirm MCAR but provides no evidence to reject
that the data MCAR. If the test has a significant p-value, it suggests
that the data is either MAR or MNAR, but its unclear which one. This
test is also vulnerable to Type 1 error (false positive result) and Type
2 error (false negative result) as the test may not detect MCAR
violations with its low statistical power, even if the MCAR assumption
holds.

## Different missing data patterns

A missing data pattern describes the pattern of missing data among the
observed data. As described by Little & Rubin (2014), it's important to
remark the missing data patterns of the data set prior to data handling
as some handling methods are intended for certain classified patterns.
The following table summarises some missing data patterns.

```{r fig.width=5, fig.height=3}
print(missing_pattern_table)
```

## Literature

While reviewing the literature on missing data, it became evident that
despite the use of established handling methods, there remains room for
improvement in how missing data are addressed in clinical study reports
and in the selection of appropriate methods.

In 2014, Powney et al. reviewed 100 longitudinal clinical trials
conducted between 2005 and 2012 and found that only 44 reported an
adequate method to handle missing data, while 30 used potentially
inadequate methods.

Spineli et al. (2015) examined 190 systematic reviews published after
2009 to investigate how missing data were reported in randomized
controlled trials. Although 175 of these reviews mentioned missing data,
only 61 discussed its implications.

Hunt et al. (2021) reviewed 62 pharmacoepidemiologic multi-database
studies from 2018 to 2019 to assess missing data reporting and handling.
Thirty-five studies reported missing data, but only 19 described methods
for handling it. The most popular approach was complete case analysis
(CCA) used in 13 studies, followed by multiple imputation (MI) in 2
studies.

Complete case analysis was also the predominant method in a systematic
review of 229 observational studies from the United Network for Organ
Sharing (UNOS) database (Baker et al., 2025). Forty-one studies used
CCA, 22 reported MI, and notably, 31 studies removed covariates due to
missing values—an action that could introduce further bias depending on
the covariates' relationship with other predictors.

\newpage

# Clinical Trial Data

We have sourced two different data sets to perform our missingness
analysis on. They both have different sample sizes and different levels
of missing data which is an advantage as we can conduct analysis in
different settings.

**Acupuncture Data**

Our first dataset contains results from a clinical trial determining the
effects of acupuncture therapy on chronic headache in primary care
(Vickers et al 2004). Vickers released the dataset from the study in
2006 and is publicly available to download in excel format from
<https://pmc.ncbi.nlm.nih.gov/articles/PMC1489946/>. The study design
for the acupuncture trial is a longitudinal randomised controlled trial
with two follow up time points. 401 participants were gathered from
general practices in England and Wales who suffered from migraines. The
main objective was to determine the effects of acupuncture use on
headache severity, health status, absence days and GP and therapist
visits in comparison to avoiding acupuncture against a control
intervention whih offered standard care. Headache and medication use was
recorded at baseline and additional factors were recorded
post-randomisation such as such sick days and daily activities. The
results of the trial showed a 34% decrease in headache severity score at
the final time point in comparison to the control group with a 16%
decrease. The acupuncture therapy group also showed a 37% decrease in
medication use, which compares to 23% in the control group.

*Data Overview*

13 variables were recorded during the acupuncture trial.

```{r fig.width=5, fig.height=3}
print(acu_var_table)
```

Variables `pk2` and `pk5` are the headache severity scores at 3 months and 12
months.

```{r fig.width=5, fig.height=3}
print(gtsummary::as_kable(acu_summary))
```

Figure 2.x is the summary statistics of the acupuncture trial data. Note
pk2 and pk5 are post randomisation variables and contain missing data.

```{r fig.width=5, fig.height=3}
print(acu_sex_plot)

```

Figure 2.x is the distribution of sex in both treatment groups of the
acupuncture trial data. The participants are predominantly female.

```{r fig.width=5, fig.height=3}
print(acu_age_plot)
```

Figure 2.x is the distribution of age in both treatment groups. Both
distributions are negatively skewed, predominantly middle aged and
older.

```{r fig.width=5, fig.height=2.5, warning=FALSE}
print(acu_pk1_plot)
```

```{r fig.width=5, fig.height=2.5, warning=FALSE}
print(acu_pk2_plot)
```

```{r fig.width=5, fig.height=2.5, warning=FALSE}
print(acu_pk5_plot)
```

Figure 2.x,2.x,2.x: These plots show the distribution of the headache
severity score at the three measured time-points. Over the three
time-points we can see a generally positively skewed distribution with
most headache severity scores ranging between 0 and 35. There appears to
be minimal difference in headache severity score over the treatment
group, and the control group appears to have a lower headache severity
score overall. The 3 month and 12 month time-point distributions are
incomplete due to the missing values.

**VITAL data**

*Source*

MacFarlane et al. conducted a study on the therapeutic effects of
vitamin D and fish oil on osteoarthritic knee pain. We sourced this data
at xyz.

*Study Design*

This study was a 2x2 factorial designed longitudinal randomised
controlled trial. 1,398 participants in the study was randomised to 4
possible treatment combinations. Participants were randomised to receive
either omega-3 oils or vitamin D, both treatments, or randomised to the
placebo group. For this research project, we assumed that that the
participants were only randomised to either receive the fish oil
treatment, vitamin D treatment or the placebo. We did not include the
interaction between vitamin D and fish oil in our substantive model or
our imputation model.

|              | **Vitamin D**            | **Placebo** |
|--------------|--------------------------|-------------|
| **Fish Oil** | N = 342 *(not analysed)* | N = 353     |
| **Placebo**  | N = 332                  | N = 371     |

*Data Overview*

There are x variables recorded in the VITAL data, the following table
refers to a subset of the variables that we frequently use.

```{r fig.align='center'}
print(vital_var_table)
```

```{r warning = FALSE}
print(gtsummary::as_kable(vital_summary))
```

Figure 2.x shows the summary statistics of the VITAL data. It provides
information on all 4 treatment groups. Note here that the VITAL data
contains missing baseline data as well as post randomisation data.

```{r fig.width=5, fig.height=3, fig.align='center', warning=FALSE}
print(age_vital_plot)
```

Figure 2.x shows the distribution of age in each treatment group, which
appear to be normal/ positive skewed distributions.

```{r fig.width=5, fig.height=3, fig.align='center', warning=FALSE}
print(bmi_vital_plot)
```

Figure 2.x shows the distribution of body mass index in each treatment
group, these appear to be positively skewed with a bmi between 20 and
40. This shows a range between healthy weighted and obese participants.

```{r fig.width=4, fig.height=3, warning=FALSE}
print(sex_vital_plot)
```

Figure 2.x is the distribution of males and females in the VITAL data.
Similar to the acupuncture trial, the participants are predominantly
female.

```{r fig.width=4, fig.height=2.5, warning=FALSE}
print(pain_base_vital_plot)
```

```{r fig.width=4, fig.height=2.5, warning=FALSE}
print(pain_yr1_vital_plot)
```

```{r fig.width=4, fig.height=2.5, warning=FALSE}
print(pain_yr2_vital_plot)
```

```{r fig.width=4, fig.height=2.5, warning=FALSE}
print(pain_yr3_vital_plot)
```

```{r fig.width=4, fig.height=2.5, warning=FALSE}
print(pain_yr4_vital_plot)
```

Figure 2.x,2.x,2.x,2.x,2.x shows the distribution of the knee pain from
baseline until 4 years post randomisation. In all treatment groups it
appears that the distributions follow a positive/ near normal
distribution and skews more positive as time goes on. This occurs in all
groups. Each time-point has an incomplete distribution due to missing
data.

\newpage

## Missing analysis

**Acupuncture Data**

The summary statistics of the acupuncture trial data shows that all
baseline variables have been observed. It appears two post-randomisation
outcome variables contain missing data. This pattern would be described
as a multivariate monotonic pattern as the percentage of missing data
increases at each follow up time point, which is common in longitudinal
studies. We can visualise this in the plot below.

```{r fig.width=5, fig.height=3}
print(acu_miss_perc_plot)
```

Figure 2.x: Proportion of missing data (dark green) in outcome variables
$pk2$ and $pk5$. This follows a monotonic missing data pattern.

If we separate the acupuncture trial data into the treatment group and
control group, we can observe that there is a higher percentage of
missing data in the control group.

```{r fig.width=5, fig.height=3}
print(acu_miss_perc_group)
```

Figure 2.x: Higher percentage of missing data in control group

If we investigate the missing data pattern in more detail, we can see
that the pattern is similar in both groups. In the plot below, we can
see that the pattern in which the three headache severity scores $pk1$,
$pk2$ and $pk5$ are observed (1,1,1) followed by the pattern in which
both post-randomised headache severity scores contain missing data
(1,0,0).

```{r fig.width=5, fig.height=3}
print(acu_miss_pattern_plot)
```

Figure 2.x: Missing data pattern is similar in both groups (1=observed,
0=missing)

**VITAL data**

The summary statistics of the VITAL data showed, unlike the acupuncture
trial data, there are also baseline variables that contain missing data
as well as the post randomised variables. A non-monotic multivariate
pattern appears to show here as there is a great proportion of missing
data in the knee pain score at the first year post randomisation, which
drastically decreases in year 2, then steadily increases again.

```{r fig.width=5, fig.height=3}
print(vital_miss_perc_plot)
```

Figure 2.x: Non-monotonic pattern across out knee pain scores in VITAL
data.

Recall that the VITAL data set contains 4 different combinations of
treatment. Separating this out in the plot below we can see the
percentage of missing data in each knee pain time point variable in each
group. The proportion appears to be similar in each group with year 2
post randomisation containing the most missing data.

```{r fig.width=5, fig.height=3}
print(vital_miss_perc_group_plot)
```

Figure 2.x: Percentage of missing data across baseline and 4 follow-up
time points. Non-monotonic pattern appears similar in 4 randomised
groups.

Looking at the possible missing data patterns in the 4 combinatory
groups, they are mostly similar with little anomalies. As this dataset
has more time-points recorded than the acupuncture trial data, and more
treatmenr groups, there are more possible missing data patterns to
observe. The most common data pattern still stands as (1,1,1,1,1) in
which data is observed at all time-points.

```{r fig.width=5, fig.height=3}
print(vital_miss_pattern)
```

Figure 2.x: Missing data patterns across 4 treatment groups in VITAL
data.

\newpage

# Methods

## Anchoring methods

We began our analysis with basic methods to serve as a comparison
benchmark for more advanced techniques. As previously mentioned, these
included complete case analysis, last observation carried forward
(LOCF), and the mean observation method. Each of these approaches was
used to anchor the results and provide a reference point for evaluating
the performance of more robust models.

### Complete case analysis

For both datasets, given that our estimand is the mean difference in pain scores at the end of the study, we fit a model regressing the final pain score on treatment group and baseline pain score. CCA can yield valid inferences when the data are MCAR, or when the outcome is MAR conditional on the variables included in the model. In our analysis, we adjust for treatment group and baseline pain score. To ensure comparability across methods and to reduce computational burden, we did not include additional covariates in any of the analyses.

We applied the following models for each dataset:

```{r, eval=FALSE, echo=TRUE}
# For acupuncture data set
acu_CAA <- lm(pk5 ~ group + pk1 , data = acu_wide)
# For VITAL data set
vital_complete <- lm(pain_yr4 ~ fishoilactive + vitdactive + pain_base, data = vital_wide)
```

One of the main drawbacks of using CCA is the loss of power due to reduced sample size, along with the risk of introducing bias. In our implementation, the lm() function in R automatically excludes any individual with missing values in the model variables. As a result, we are left with only 301 out of 401 observations in the acupuncture dataset and 698 out of 1390 in the VITAL dataset—corresponding to a loss of approximately 25% and 50% of the sample size, respectively.

We have not included many covariates in our models, so bias could exist even with complete data due to omitted variables. Now, suppose we attempted to adjust for all relevant baseline covariates—missing data would still pose a problem. For instance, in the VITAL dataset, several baseline covariates have missing values. Excluding observations with missing data in these variables could still induce bias, particularly if the missingness is related to the outcome.

In the acupuncture dataset, although there is no missingness in baseline variables, CCA has further limitations. While intermediate outcomes such as the 3-month pain score or post-randomisation variables like pain frequency may contain valuable information, they are not included in our models. This is not due to a limitation of CCA or LME itself, but because including post-randomisation variables would violate the goal of estimating the total effect of randomisation. However, these variables can still contribute indirectly through MI. Rather than adjusting for them in the model, MI treats repeated and related measures as part of a multivariate distribution within each individual, allowing partially observed variables to help impute missing values without compromising the target estimand.

In rare cases—such as when the sole interest is in estimating the mean
difference between groups at the end of the study, and there are no
important non-baseline covariates complete case analysis can still yield
valid results. Thus when apply complete case analysis, we are assuming:
the outcome is Missing At Random (MAR), conditional on the observed
baseline predictors included in the model.

This condition sometimes referred to as "conditionally MCAR". While it
is weaker than the strict MCAR assumption, it is still quite strong. It
implies that:

-   Missingness depends only on the variables included in the model;
-   The model is correctly specified;
-   The outcome variable is not involved in the missingness
    mechanism—consistent with the MAR framework.

Under these assumptions, CCA can provide unbiased estimates. However,
they are rarely fully met in practice, especially in complex
longitudinal studies where many relevant variables and time points are
involved.

This can be expressed in the mathematics form, assuming outcome of
interest Y MAR depends on baseline covaraites X

\begin{align*}
  Pr(R_i=1 \mid Y_i, X_i) = Pr(R_i=1 \mid X_i)
  \end{align*}

Thus, we can infer under MAR, the distribution of outcome within
covaraites is the same in the observed data, the unobserved data, and
the population.

\begin{align*}
  & Pr(Y_i \mid R_i=1, X_i) \\ 
  & = \displaystyle \frac{Pr(R_i=r,Y_i,X_i)}{Pr(R_i=1,X_i)} \\
  & = \displaystyle \frac{Pr(R_i=1 \mid Y_i,X_i)Pr(Y_i,X_i)}{Pr(R_i=1 \mid X_i)Pr(X_i)} 
  \\
  & Assuming \ \ \Pr(R_i=1 \mid Y_i, X_i) = Pr(R_i=1 \mid X_i) \\
  & = \displaystyle \frac{Pr(Y_i,X_i)}{Pr(X_i)} \\
  & = Pr(Y_i \mid X_i)
  \end{align*}

The key issue is that we aim to include as much relevant information as
possible in the set of predictors X to make the MAR assumption more
plausible and robust. As discussed above, more statistically principled
methods—such as multiple imputation or mixed-effects models—allow us to
incorporate a broader set of variables and handle missingness more
effectively. However, it is important to note that the MAR assumption is
fundamentally untestable. Therefore, to assess the robustness of our
conclusions, we must perform sensitivity analyses, which are presented
in later sections.

### Last observation carry forward

Last Observation Carried Forward (LOCF) is a simple imputation method
that fills in missing values based on the last observed outcome for each
subject. It relies on assumptions unrelated to the missing data
mechanism, meaning it does not model the reason for missingness.
Instead, LOCF assumes that the outcome remains stable after dropout,
which may be plausible in some clinical contexts, such as when the
treatment effect has plateaued or when patients are expected to remain
in a steady state.

However, this assumption is not convincing in either of the datasets
analyzed here. In both the Acupuncture and VITAL trials, pain score are
unlikely to be stabilized throughout, and the outcomes display trends
over time, making the LOCF assumption potentially misleading and
unsuitable for accurate estimation of treatment effects.

### Mean observation method

Mean imputation is another simple method that fills in missing values by
replacing them with the mean of the observed data at each timepoint.
Like LOCF, it is based on assumptions unrelated to the missing data
mechanism and does not attempt to model why the data are missing.

This method implicitly assumes that the mean of the observed values
accurately represents the population mean, which can be problematic. In
fact, this can be viewed as an even more radical assumption than
assuming data are Missing Completely At Random (MCAR), since it ignores
potential bias introduced by the missingness and oversimplifies
individual variation.

As with LOCF, this assumption is not convincing in either of the
datasets analyzed here. In both the Acupuncture and VITAL trials, pain
scores vary over time, and individual trajectories differ. Replacing
missing values with the average observed value at each timepoint
disregards this variation and may lead to underestimated variability and
biased treatment effect estimates.

## Changing imputation methods

To address missing data in a statistically principled manner, it is helpful to distinguish between two components: the imputation step and the modeling step. In our anchoring methods, we either did not perform imputation—as in CCA—or relied on single imputation techniques, such as LOCF or mean imputation.

While methods like LME can utilize most available observed data and yield unbiased estimates under the MAR assumption without imputation, they cannot incorporate other post-randomisation variables that are not part of the model (e.g., pain frequency in the acupuncture dataset). In contrast, MI allows the inclusion of such auxiliary information during the imputation process.

In most realistic scenarios, doing no imputation or applying single imputation methods is likely to introduce bias and underestimate uncertainty, especially when missingness is related to unobserved outcomes. Thus, more flexible approaches—such as MI combined with appropriate modeling—are generally preferred for robust inference in the presence of missing data.

### Multiple imputation

A key limitation of single imputation methods is that they treat imputed
values as if they were observed data when fitting the substantive model.
This approach fails to reflect the inherent uncertainty associated with
missing values—it assumes we have perfectly recovered the unobserved
data, which is never the case in practice.

In reality, the best we can do is to estimate the distribution of the
missing data conditional on the observed data, under specific
assumptions about the missingness mechanism (e.g., MAR). Importantly,
this distribution is not adequately captured by a single draw, as is
done in single imputation. Instead, to account for uncertainty, we must
generate multiple draws from this distribution, resulting in multiple
imputed datasets. This is the foundation of Multiple Imputation (MI).

As we will demonstrate in the following sections, the draws used in MI
must be (at least approximately) Bayesian for Rubin’s variance formula
to yield valid inference. By fitting the substantive model separately to
each of the K completed datasets, we obtain multiple estimates that,
when pooled, incorporate both within-imputation variability and
between-imputation variability. This approach not only addresses bias
caused by missing data but also appropriately inflates standard errors
to reflect the uncertainty introduced by imputation. Rubin’s rules
provide a general framework for combining these estimates to obtain
point estimates, variances, confidence intervals, and statistical tests.

We also briefly introduce the different methods available for performing
multiple imputation. In cases with a monotonic missing data pattern,
sequential regression offers a straightforward solution. For arbitrary
or multivariate missingness, more flexible approaches like joint
modeling or Fully Conditional Specification (FCS)—also known as multiple
imputation by chained equations (MICE)—are required.

In our project, we chose the FCS approach, given its flexibility and
ease of use when dealing with datasets like Acupuncture and VITAL that
have complex missingness patterns across multiple variables. A brief
comparison of joint modeling versus FCS is included below to justify
this decision.

While we used five imputations (K = 5) for our MI procedure in this
analysis, we also discuss the considerations involved in choosing the
number of imputations. These include factors such as the proportion of
missing data, computational cost, and the desired precision of standard
errors and confidence intervals.

#### Rubin's rules

Once multiple imputed datasets are generated, Rubin’s rules are used to
combine the parameter estimates and associated uncertainty across the K
imputed datasets. The steps are as follows:

1.  Impute K complete datasets, each containing different plausible
    values for the missing data. (More details in later section, we will
    focus on the following steps for now)

2.  Fit the substantive model (e.g., a linear model or mixed-effects
    model) to each imputed dataset, obtaining Parameter estimate
    $\hat{\beta}_{k}$ and Variance estimate $\hat{\sigma}_{k}^{2}$ for
    each k=1,2,...,K

3.  Compute the pooled estimate $\hat{\beta}_{MI}$ and its total
    variance $\hat{V}_{MI}$: \begin{align*}
      \hat{\beta}_{MI} = \frac{1}{K} \sum_{k=1}^{K}{\hat{\beta}_{k}} \\
      \hat{V}_{MI} = \hat{W} + (1 + \frac{1}{K}) \hat{B} \\
      \hat{W} = \frac{1}{K} \sum^{K}_{k=1}{\hat{\sigma}^{2}_{k}} \\
      \hat{B} = \frac{1}{K-1} \sum^{K}_{k=1}({\hat{\beta}_{k}} - \hat{\beta_{MI}})^{2} \\
      \end{align*}

4.  To test a null hypothesis $\hat{\beta}_{MI} = \beta^{0}$ use a
    t-statistic with $\nu$ degrees of freedom. This allows us to
    construct confidence intervals and perform hypothesis testing,
    reflecting the additional uncertainty due to imputation.
    \begin{align*}
      T = \frac{\hat{\beta}_{MI} - \beta^{0}}  {\sqrt{\hat{V}_{MI}}} \\
      \nu = (K-1)[1 + \frac{\hat{W}}{(1 + 1/K) \hat{B}}]^{2}
      \end{align*}

#### Sequential regression MI

In many longitudinal studies, the missing data pattern is approximately
monotonic, particularly when dropout is due to participant withdrawal, a
common situation in clinical research. In such cases, later measurements
are often missing while earlier ones are observed, which permits the use
of sequential regression for imputation under the assumption of Missing
At Random (MAR).

To justify this, consider the joint distribution of the outcome vector
for individual $i$ can be factorized as:

\begin{align*}
    f(Y_{i,1},Y_{i,2},...,Y_{i,p}) = f(Y_{i,p} \mid Y_{i,1},...,Y_{i,p-1}) * 
    f(Y_{i,p-1} \mid Y_{i,1},...,Y_{i,p-2}) * ... * f(Y_{i,2} \mid Y_{i,1}) 
    * f(Y_{i,1})
    \end{align*}

Under a monotonic missingness pattern, for each missing value $Y_{i,j}$
all preceding values $Y_{i,1},...,Y_{i,j-1}$ are observed. If we also
assume Missing At Random (MAR), then each of these conditional
distributions can be estimated directly from the observed data. This
provides a principled basis for sequential regression imputation, where
each variable is regressed on the variables preceding it in order, and
missing values are imputed based on those conditional models.

Suppose we have $i=1,...,n$ individual with $j=1,...,p$ variables. When
the missing data pattern is monotonic, sequential regression provides an
efficient and valid approach to imputation under the MAR assumption. The
procedure follows these steps:

1.  specify the model

\begin{align*}    
    Y_{i,j} = (1,Y_{i,1},...,Y_{i,j-1})^{T} * \beta_{j} + 
    e_{i,j},\ {e_{i,j}^{i.i.d.} \sim{N} (0,\sigma^{2}_{j})}
    \end{align*}

2.  Under the monotonic pattern, for every missing $Y_{i,j}$ we assume
    that $Y_{i,1},...,Y_{i,j-1}$ are fully observed. We fit the
    regression model using ordinary least squares (OLS) to obtain
    estimates $\hat{\beta}_{j},\hat{\sigma}_{j}^{2}$

3.  To incorporate uncertainty, we draw new parameters
    $\beta_{j}^{*},\sigma_{j}^{*2}$ from their posterior distributions:

\begin{align*}
   \sigma_j^{*2} = \frac{\hat{\sigma}_j^2 (n_j - j)}{z} \\
   \beta^* \sim N(\hat{\beta}, \sigma_j^{*2} A_j) \\
   A_j = (\sum_{i=1}^{n_j} x_{i,j} x_{i,j}^T)^{-1}
   \end{align*}

4.  For individuals with missing $Y_{i,j}$ generate imputations from the
    model:
    $Y_{i,j} = (1,Y_{i,1},...,Y_{i,j-1}) \beta^{*} + e^{*}_{i,j}$,
    $e_{i,j}^{*} \sim{N} (0,\sigma^{*2}_{j})$

5.  Repeat for $Y_{i,j+1}$ until complete

This sequential regression method is applicable to datasets with a
monotonic missingness pattern, such as the Acupuncture dataset, where
individuals tend to drop out in a consistent, time-ordered manner.
However, it is not suitable for datasets with non-monotonic missingness,
such as VITAL, where some individuals may have missing values at
intermediate time points but return for later follow-ups. In such cases,
more flexible approaches like Joint Modeling or Fully Conditional
Specification (FCS) are required to properly handle the complex,
arbitrary missing data structure.

#### Joint modelling

When the missing data pattern is non-monotonic, as in the VITAL dataset,
sequential regression is no longer appropriate. Instead, we can apply
joint modeling, which makes no assumption about the missingness pattern
but assumes the missing data mechanism is ignorable (typically, Missing
At Random).

Under joint modeling, we assume that the complete multivariate outcome
vector follows a multivariate normal distribution: \begin{align*}
    & Y \sim N(\beta,\ \Omega) \\
    & Y = (Y_{i,1},Y_{i,2},...,Y_{i,p})^{T} \\
    & \beta = (\beta_{0,1},\beta_{0,2},...,\beta_{0,p})^{T} \\
    & \Omega \ \text{is the covariance matrix}
    \end{align*}

To estimate the parameters $\beta$ and $\Omega$ and impute missing
values, Gibbs sampling is one of the approaches. This algorithm draws
each parameter in turn, conditional on the current values of all other
parameters and the data.

To get priors to start Gibbs sampling. We begin with initial estimates
$\beta^{0}$ and $\Omega^{0}$, computed from the observed data. For each
variable with missing values, we also generate an initial imputation
$Y_{M}^{0}$ by sampling from the observed values of that variable with
replacement. This allows us to calculate initial statistics such as
$\overline{Y}^{0}$ and and the sample covariance matrix $S^{0}$ (Note it
also used as prior sample covariance matrix $S^{P}$ in each iteration)

Then, for each iteration r the following steps are performed:

1.  Draw the precision matrix (inverse
    covariance):$\Omega^{-1,r} \sim W(n + \nu, (S_p^{-1} + S^{r-1})^{-1})$

2.  Draw the mean
    vector:$\beta^{r} \sim N(\bar{Y}^{r-1}, n^{-1} \Omega^r)$

3.  Impute missing values:
    $Y_{M}^{r} \sim f(Y_{M} \mid \beta^{r}, \Omega^{r}, Y_{O})$

4.  Update the mean and covariance estimates: $\bar{Y}^{r}$ the mean of
    the combined imputed and observed data, and $S^{r}$ the sum of
    squares and cross-products from the combined data

After a sufficient number of burn-in iterations, we repeat this process
K times to generate K imputed datasets. These are then analyzed using
the substantive model of interest, and Rubin’s rules are applied to pool
the results, yielding valid parameter estimates and standard errors that
account for the uncertainty due to missing data.

#### Full conditional specification

Fully Conditional Specification (FCS), also known as multiple imputation
by chained equations (MICE), is an extension of sequential regression
imputation that relaxes the requirement that all covariate values used
in the regressions be fully observed.

Importantly, when the missingness pattern is monotonic, FCS becomes
equivalent to the sequential regression method discussed earlier.
However, its key advantage is that it remains valid under non-monotonic
missingness, making it suitable for more general data structures like
the VITAL dataset.

The term “full conditional specification” refers to the fact that each
variable is imputed from its full conditional distribution, given all
other variables. This allows for more flexible modeling of multivariate
missingness.

The general procedure involves the following steps:

1.  Reorder the variables so that the overall missingness pattern is as
    close to monotonic as possible. This can improve stability and
    convergence in the imputation process.

2.  Initialize missing values by filling in initial guesses—often by
    drawing, with replacement, from the observed values of each
    variable.

3.  For each variable $Y_{j}$

    -   Regress the observed part of $Y_{j}$ on all other variables
        (including those with imputed values).
    -   Use the fitted model to impute the missing values in $Y_{j}$,
        treating the other variables as given.

4.  Repeat step 3 for all variables with missing data to complete one
    cycle.

5.  Perform multiple cycles until convergence, and then repeat the
    entire process K times to generate K imputed datasets.

These datasets can then be analyzed with the substantive model, and the
results pooled using Rubin’s rules. FCS is widely used in practice due
to its flexibility and implementation in tools like the mice package in
R.

#### FCS VS joint modelling

Generally, under the assumption of a multivariate normal distribution,
the joint distribution uniquely determines the full set of conditional
distributions, and vice versa. This means that, in theory, joint
modeling and fully conditional specification (FCS) are mathematically
compatible representations of the same underlying structure—provided all
models are correctly specified.

In practice, however, joint modeling using a Gibbs sampler is often
considered a more efficient algorithm. It also has the advantage of
allowing the inclusion of prior information, which can be particularly
useful when data are sparse or when integrating external knowledge into
the model. Additionally, joint modeling methods can incorporate ridge
parameters to stabilize the estimation of the covariance matrix, which
becomes important when the number of variables is large relative to the
sample size.

However, these concerns do not apply in our project, as our datasets
have moderate dimensionality and sufficient sample size. Therefore, we
opted to use Fully Conditional Specification (FCS) for multiple
imputation.

We adopted the FCS approach using the mice package in R to perform multiple imputation throughout this project. One advantage of FCS is its ease of implementation, as it does not require explicit specification of a joint model or prior distributions, and it typically converges in fewer iterations in practice. Moreover, the balanced longitudinal design of both datasets—where follow-up measurements occur at consistent time points across individuals—makes FCS especially suitable. This structure allows for sequential imputation of partially observed variables in a consistent order across patients. In contrast, if the data were collected at irregular or patient-specific time points, or under a more complex design such as a cluster randomised trial, the FCS approach would be more difficult to apply reliably.

Finally, within the MICE framework, different imputation methods can be
specified for different variable types—such as predictive mean matching
or sampling from observed values. We will explore the impact of these
options later in the analysis.

#### How to choose imputation number K

Another practical consideration in multiple imputation (MI) is the
choice of the number of imputations, denoted by K. While early
applications often used K=5, more recent work emphasizes that the
optimal number depends on the degree of missing information in the data.

A key parameter in determining K is $\gamma_{0}$, which represents the
fraction of missing information for the parameter of interest.
Unfortunately, $\gamma_{0}$ is typically unknown in advance, including
in our project.

To address this, Bodner (2008) proposed a simple and conservative
strategy: using the proportion of complete cases in the dataset as a
proxy for $1-\gamma_{0}$, thereby estimating $\gamma_{0}$
conservatively. This approach allows for an informed yet practical
choice of K, especially when precise calculation of the missing
information is not feasible.

In our analysis, we used K=5 imputations as a baseline, while
recognizing that more imputations may be needed in cases with higher
levels of missingness or if more precise estimates of standard errors
are required. We will explore the implication of changing K in the
following sessions.

**Choose 3-5 imputations**

The classic advice for multiple imputation is to use a low number of
imputations, typically between 3 and 5, when the proportion of missing
information is moderate. As discussed in Rubin (1987), the argument for
choosing a small K is based on the total variance
estimate:$T_{K} = (1 + \frac{\gamma0}{K}) T_{\infty}$

where $T_{K}$ is the variance with K imputations, $T_{\infty}$ is the
asymptotic variance as $K \rightarrow \infty$ and $\gamma_{0}$ is the
fraction of missing information. Since $\gamma_{0}$ is typically
unknown, this formula helps illustrate the trade-off between the number
of imputations and efficiency.

There is often limited benefit in increasing K beyond 5. For instance,
if $\gamma0=30\%$, using K = 5 result in $T_m=1.06T_\infty$, indicating
only a 6% inflation in variance compared to the ideal case.

In this project, we chose to use K=5 imputations for multiple
imputation, following the classical recommendation to use a low number
of imputations when the proportion of missing information is moderate.

**Chosse \>20 imputations**

While early guidelines recommended using a low number of imputations
(typically K=3-5) for moderate missingness, more recent research argue
that increasing K beyond this range can yield important gains in
statistical efficiency.

-   Royston (2004) suggested that to constrain the coefficient of
    variation of $ln(t_{\nu}\sqrt{T})$ to below 0.05—effectively keeping
    the width of confidence intervals within about 10% uncertainty, a
    minimum of K\>20 is required. Here, $t_{\nu}$ is the quantile from a 
    t-distribution with ${\nu}$ degrees of freedom, reflecting the uncertainty
    from finite imputations, and $T$ is the total variance combining within- 
    and between-imputation components.
-   Graham (2007) argued that to achieve statistical power within 1% of
    the theoretical maximum, researchers should use at least K=20.
-   Bodner (2008) examined how the number of imputations relates to the
    fraction of missing information $\gamma_{0}$, and its effect on
    p-values and confidence intervals. He recommended increasing
    $K=(3,6,12,24,59,114,258)$ for $\gamma0=(0.1, 0.3, 0.5, 0.7, 0.9)$
    accordingly.

In some situations—such as when estimating variance components or when
dealing with highly uncertain estimands—using a very high number of
imputations (e.g.,K=200) may be warranted to approximate the full
posterior distribution.

The main drawback of increasing K is that it leads to longer
computational time. However, this is generally the only limitation, and
it becomes manageable with modern computing resources. Moreover,
starting with a high number of imputations gives greater flexibility: we
can always test the stability or sensitivity of our results by
re-analyzing a subset of the imputations (e.g., comparing the
performance at K=5,10,20) without needing to re-run the entire
imputation process.

Thus, while K=3-5 is often sufficient under moderate missingness and
when the focus is on point estimates, using a larger K can improve
robustness in more demanding settings, with minimal trade-offs beyond
processing time.

$K \approx 100\lambda$

A widely cited rule of thumb proposed by White, Royston, and Wood (2011)
recommends choosing the number of imputations based on the fraction of
incomplete cases in the dataset, denoted as $\lambda$. Specifically,
they suggest setting:$K \approx 100\lambda$

This rule has become a de facto standard, particularly in medical
research, due to its simplicity and strong theoretical support. The key
idea is that the number of imputations should roughly match the
percentage of individuals with any missing data.

-   The Monte Carlo error of the pooled point estimate $\hat{\beta}$ is
    approximately 10% of its standard error.

-   The Monte Carlo error of the test statistic
    $\hat{\beta}/SE_{\hat{\beta}}$ is roughly 0.1.

-   The Monte Carlo error of a p-value is approximately 0.01 when the
    true p-value is 0.05.

These error bounds are typically acceptable in applied research,
ensuring stable estimates and valid inference without requiring an
excessive number of imputations.

One challenge with applying this rule arises in high-dimensional
settings, where the number of variables is large. In such cases, it is
common for a large proportion of individuals to have at least one
missing value, which can push $\lambda$ close to 1. To address this, it
is reasonable to use the overall missing rate (i.e., total proportion of
missing cells in the dataset) as a conservative proxy for $\lambda$ when
needed.

This rule provides a useful upper bound for choosing K, especially when
balancing the goals of statistical precision and computational
efficiency.

## Changing substantive model

Another important decision point in the missing data handling process is
the choice of the substantive model—the model used for the final
analysis after imputation. One natural alternative to standard multiple
linear regression is the linear mixed-effects model (LME).

As discussed earlier, LME has the advantage of leveraging all available
observed data, even in the presence of missingness. In fact, under
certain conditions, LME can yield valid inferences without requiring
multiple imputation, as long as the missingness mechanism is ignorable
(e.g., MAR) and auxiliary variables are not essential.

Moreover, even with fully observed data—either originally complete or
completed through imputation—LME remains a superior choice in many
longitudinal settings because it explicitly accounts for within-subject
correlation from repeated measurements. This leads to more accurate
estimates and better statistical efficiency than standard linear models,
which ignore the data’s hierarchical structure.

In this project, we focus on multiple linear regression and LME, both of
which assume a linear relationship between covariates and the outcome.
However, when selecting a substantive model, it's important to evaluate
this assumption. In cases where linearity is questionable, alternative
models—such as polynomial regression or spline-based models—may provide
a better fit and should be considered.

An additional benefit of using LME is that it allows for flexible
modeling of time. Specifically, it enables us to treat time as a
continuous variable, rather than as a categorical factor, which can
improve interpretability and statistical power. This modeling choice
will be explored further in a later section.

## Data wrangling

The VITAL dataset was originally in wide format, which is not suitable
for methods like linear mixed-effects models (LME) that require
long-format data. To address this, the data was reshaped to long format

In the case of multiple imputation (MI), imputation was first performed
on the wide-format data using the mice() function. The resulting imputed
datasets were then combined and transformed to long format, allowing for
the inclusion of time as a continuous variable. This approach enables
the application of LME as the substantive model, which effectively
accounts for repeated measurements over time. Moreover, it accommodates
more flexible timing of follow-up assessments.

One of the challenge when we want to do MI and then fit into LME as substantiv model is that. We can only fit long format into LME, but doing imputation with long format will treat each observation as individual which is incorrect. So we use the wide format to do MI first, then change to LME. Below is a simple schema to illustrate the process (Outcome1/2/3 indicates outcome at differnt time point, and x indicates missing data).

Data in wide format:

|ID|Outcome1|Outcome2|Outcome3|
|---|---|---|---|
|1001|2.5|3.5|x|
|1002|4.5|x|6.3|
|1003|3.3|6.2|8.1|

Completed data after MI in wide format:

|ID|Outcome1|Outcome2|Outcome3|
|---|---|---|---|
|1001|2.5|3.5|4.5|
|1002|4.5|5.2|6.3|
|1003|3.3|6.2|8.1|

Transfored completed data in long format:

|DI|Time|Outcome|
|---|---|---|
|1001|1|2.5|
|1001|2|3.5|
|1001|3|4.5|
|1002|1|4.5|
|1002|2|5.2|
|1002|3|6.3|
|1003|1|3.3|
|1003|2|6.2|
|1003|3|8.1|

Comments

Why just refer to the VITAL dataset and how the original format here? 

You should raise the general point that the data can be represented in both wide and long format for both studies, and you will need to interchange between the two depending on the substantive model and the missing data method i.e. imputation method to be used.

It's also crucial to highlight the longitudinal design of both studies is balanced, i.e. there was a planned, equal number of follow-up measurements at roughly fixed time points. This will make these studies amenable to a more straightforward imputation of partially observed follow-up variables as part of a multivariate distribution.

In our project, both dataset have balanced longitudianal design. Which is not always the case and could complex the analysis. This make our analysis 
a more straightforward imputation of partially observed follow-up variables as part of a multivariate distribution.

## Data Wrangling

Both the Acupuncture and VITAL datasets originate in wide format, where each subject's repeated measurements are stored in separate columns. While this format is convenient for some operations, it is not directly compatible with LME, which require the data to be transformed into long format—with one row per observation per time point.

In our project, both datasets have a balanced longitudinal design, with follow-up measurements taken at regular, pre-specified intervals. This is not always the case in real-world studies, where irregular or missing time points can complicate analysis. The balanced structure in our data allows for a more straightforward imputation of partially observed follow-up outcomes using a multivariate distribution, making the MI process more reliable and easier to implement.

To address these differing requirements, the following general approach was used for both data sets:

1. Perform MI in wide format using the `mice()` function, treating the repeated measures as separate but related variables.
2. Convert the imputed wide data sets to long format, which allows time to be included as a continuous variable in the LME model.
3. Fit LME models on the long-format data to model outcome trajectories over time while appropriately accounting for missing values and repeated measures.

This strategy allows us to use multiple imputation methods compatible with the multivariate normal assumption across time points, while still leveraging the advantages of LME models in the analysis stage.

The balanced longitudinal design of both studies further supports this approach: each subject was planned to have the same number of repeated measurements at roughly equal intervals (monthly in the Acupuncture study and yearly in the VITAL study). This regular measurement structure allows partially observed follow-up outcomes to be imputed more reliably using a multivariate framework.

Below is a simple schematic to illustrate the reshaping process. Here, `Outcome1/2/3` represent repeated outcomes across three timepoints, and `x` indicates a missing value:

**Original data (wide format):**

| ID   | Outcome1 | Outcome2 | Outcome3 |
|------|----------|----------|----------|
| 1001 | 2.5      | 3.5      | x        |
| 1002 | 4.5      | x        | 6.3      |
| 1003 | 3.3      | 6.2      | 8.1      |

**Completed data after multiple imputation (still wide format):**

| ID   | Outcome1 | Outcome2 | Outcome3 |
|------|----------|----------|----------|
| 1001 | 2.5      | 3.5      | 4.5      |
| 1002 | 4.5      | 5.2      | 6.3      |
| 1003 | 3.3      | 6.2      | 8.1      |

**Transformed data (long format):**

| ID   | Time | Outcome |
|------|------|---------|
| 1001 | 1    | 2.5     |
| 1001 | 2    | 3.5     |
| 1001 | 3    | 4.5     |
| 1002 | 1    | 4.5     |
| 1002 | 2    | 5.2     |
| 1002 | 3    | 6.3     |
| 1003 | 1    | 3.3     |
| 1003 | 2    | 6.2     |
| 1003 | 3    | 8.1     |

## Examing using forest plot

To visually compare how our estimand (treatment effect) changes across
different missing data methods, we present results using a series of
forest plots. These plots allow us to directly assess the impact of each
imputation or modeling strategy on the estimated effect and its
confidence interval.

The same process was applied to both the Acupuncture and VITAL datasets.
For the VITAL trial, which follows a 2×2 factorial design, we simplify
the analysis—as previously discussed—by treating fish oil and vitamin D
as if they were tested in two separate randomized controlled trials.
This means we ignore potential interaction effects between the two
treatments and estimate their effects independently, which allows for a
more straightforward comparison across methods.

### Change imputation method and substantive model

We evaluated the impact of missing data by systematically varying both
the imputation method and the substantive model used in the analysis. By
doing that, we try to isolate the effects of changing either the
imputation strategy or the analysis model, and highlights the practical
impact of each decision on the resulting treatment effect estimates.

First, as described earlier, we applied three anchoring methods:
Complete Case Analysis (CAA), Last Observation Carried Forward (LOCF),
and mean imputation. Each was followed by multiple linear regression to
estimate the treatment effect.

Next, we performed Multiple Imputation (MI) using the predictive mean
matching method with 5 imputations, again analyzing the imputed datasets
using multiple linear regression—maintaining consistency with the
anchoring models for fair comparison.

Then, we changed the substantive model to a linear mixed-effects model
(LME), without performing any imputation. This approach uses all
available observed data and accounts for repeated measurements, offering
a more efficient use of the data under the MAR assumption.

Finally, we combined both improvements: we performed MI using predictive
mean matching with 5 imputations, and analyzed the completed datasets
using the LME model.

### Change estimand

Assuming our primary objective remains to estimate the mean difference in pain scores at the final timepoint for both datasets, we now explore how this estimand can be more efficiently estimated by incorporating interim outcome data.

In the previous section, we applied six different methods, two of which used Linear Mixed-Effects (LME) models as the substantive model. However, in those implementations, time was treated as a categorical variable. This approach limited our ability to take full advantage of the repeated measures structure, as it did not model the underlying trajectory of pain over time.

To address this, we revise the estimand by treating time as a continuous variable. This allows us to model individual pain trajectories across the study period, thereby leveraging all available interim outcome data. As a result, we improve the precision of the estimated treatment effect at the final timepoint—even in the presence of missing data.

#### Original Estimand - Categoricl Time

*Acupuncture data set*

The goal of the original acupuncture study was to estimate the effect of acupuncture therapy versus general care on chronic headache severity. The estimand of interest is the mean difference in headache scores at 12 months, conditional on baseline severity. This can be expressed with the following linear regression model:

$$pk5_i = \beta_0 + \beta_1 group_i + \beta_2 pk1_i + \varepsilon_i$$
where

-   $pk5_i$: headache pain score of individual $i$ at 12 months.
-   $pk1_i$: adjusted baseline headache pain score of individual $i$
-   $group_i$: binary indicator of treatment assignment (1 = acupuncture, 0 = control)
-   $\beta_0$: intercept, representing the expected 12-month pain score for a control participant with zero baseline pain
-   $\beta_1$: treatment effect of acupuncture relative to control; this is the estimand of interest
-   $\beta_2$: effect of baseline headache severity on the 12-month outcome
-   $\varepsilon_i$: residual error term for individual $i$

*VITAL data set*

In the VITAL study, the primary aim was to evaluate the effects of vitamin D and fish oil supplementation on knee pain four years after randomisation. The estimand of interest is the mean difference in knee pain scores at 4 years, conditional on baseline pain. This is modelled using the following linear regression:

$$painyear4_i = \beta_0 + \beta_1 fishoilactive_i + \beta_2 vitdactive_i + \beta_3 painbase_i + \varepsilon_i$$

where

-   $pain_yr4_i$: knee pain score for individual $i$ at 4 years post-randomization
-   $fishoilactive_i$: binary indicator for fish oil treatment (1 = active, 0 = placebo)
-   $vitdactive_i$: binary indicator for vitamin D treatment (1 = active, 0 = placebo)
-   $painbase_i$: baseline knee pain score for individual $i$
-   $\beta_0$: intercept, representing the expected 4-year pain score for a participant receiving neither treatment and with zero baseline pain
-   $\beta_1$: effect of fish oil supplementation; this is the estimand of interest in the fish oil analysis
-   $\beta_2$: effect of vitamin D supplementation; this is the estimand of interest in the vitamin D analysis
-   $\beta_3$: effect of baseline pain on the 4-year outcome
-   $\varepsilon$: residual error term for individual $i$

#### Revised Estimand – Continuous Time

To better utilise repeated measures in both datasets, we revise the estimand using a Linear Mixed-Effects (LME) model with time treated as a continuous variable.

To ensure that the intercept represents the estimated treatment effect specifically at the end of follow-up, we re-center the time variable:
- In the acupuncture study, we define $time_c = time - 12$ (in months), so that time zero corresponds to month 12, which is the end of study.
- In the VITAL study, we define $time_{contin} = time - 4$ (in years), so that time zero corresponds to year 4, which is the end of study.

With this centering, the model intercept represents the mean pain score at the end of the study, and the coefficient for the treatment group becomes the estimand of interest.

*Acupuncture data set*

The revised LME model for the acupuncture dataset is specified as:

$$
pk5_{ij} = \beta_0 + \beta_1 group_i + \beta_2 time_{c,ij} + \beta_3 (group_i \times time_{c,ij}) + \beta_4 pk1_i + b_{0i} + b_{1i} time_{c,ij} + \varepsilon_{ij}
$$

where

- $pk5_{ij}$: headache pain score for individual $i$ at time $j$
- $group_i$: binary indicator of treatment assignment (1 = acupuncture, 0 = control)
- $time_{c,ij}$: time in months since baseline, centered at 12 months ($time_c = time - 12$)
- $pk1_i$: adjusted baseline headache pain score of individual $i$
- $\beta_0$: intercept, representing the expected pain score at 12 months (end of study) for the control group with zero baseline pain
- $\beta_1$: treatment effect of acupuncture relative to control at 12 months; this is the estimand of interest
- $\beta_2$: average rate of change in pain over time in the control group
- $\beta_3$: treatment-time interaction effect
- $\beta_4$: effect of baseline headache pain
- $b_{0i}$: random intercept for individual $i$
- $b_{1i}$: random slope for time for individual $i$
- $\varepsilon_{ij}$: residual error term for individual $i$ at time $j$

*VITAL data set*

The revised LME model for the VITAL dataset is specified as:

$$
pain_{ij} = \beta_0 + \beta_1 fishoilactive_i + \beta_2 vitdactive_i + \beta_3 time_{contin,ij} + \beta_4 (fishoilactive_i \times time_{contin,ij}) + \beta_5 (vitdactive_i \times time_{contin,ij}) + \beta_6 painbase_i + b_{0i} + b_{1i} time_{contin,ij} + \varepsilon_{ij}
$$

where

- $pain_{ij}$: knee pain score for individual $i$ at time $j$
- $fishoilactive_i$: binary indicator for fish oil treatment (1 = active, 0 = placebo)
- $vitdactive_i$: binary indicator for vitamin D treatment (1 = active, 0 = placebo)
- $time_{contin,ij}$: time in years since baseline, centered at 4 years ($time_{contin} = time - 4$)
- $painbase_i$: baseline knee pain score for individual $i$
- $\beta_0$: intercept, representing the expected pain score at 4 years (end of study) for a participant receiving neither treatment and with zero baseline pain
- $\beta_1$: effect of fish oil at 4 years; this is the estimand of interest in the fish oil analysis
- $\beta_2$: effect of vitamin D at 4 years; this is the estimand of interest in the vitamin D analysis
- $\beta_3$: average rate of change in pain over time in the placebo group
- $\beta_4$: fish oil-time interaction effect
- $\beta_5$: vitamin D-time interaction effect
- $\beta_6$: effect of baseline knee pain
- $b_{0i}$: random intercept for individual $i$
- $b_{1i}$: random slope for time for individual $i$
- $\varepsilon_{ij}$: residual error term for individual $i$ at time $j$



### Change FSC methods

Even within the FCS (Fully Conditional Specification) framework, there
are multiple imputation strategies we can explore. We begin with
deterministic prediction (method="norm.predict"), which imputes missing
values using predicted values from a linear regression model. Since no
uncertainty is added, this is not technically multiple imputation—each
missing value is replaced by a single fixed estimate.

To account for uncertainty in the outcome, we adds random noise to the
regression prediction (method="norm.nob"). Further, to reflect
uncertainty in the model parameters, norm draws both regression
coefficients and residual variance from their posterior distributions,
providing a fully Bayesian imputation (method="norm").

Alternatively, instead of model-based predictions, we can impute using
observed values. predictive mean matching (method="pmm") selects donor
values from observed data that are closest in predicted value to the
missing case. Weighted predictive mean matching (method="midastouch")
extends this by weighting the distance between predicted values,
offering a more robust variant of PMM.

Or even, if we prefer a non-parametric imputation approach, we can use
the methods like random sampling (method="sample"), which imputes
missing values by randomly sampling from the observed values of the same
variable. This method does not rely on any model or predictor variables,
and instead preserves the marginal distribution of the variable.

### Change imputation numbers

As discussed above, there are various arguments for choosing a higher
number of imputations based on the fraction of missing information and
the desired precision of inference. To explore the impact of increasing
K, we adjusted the number of imputations accordingly in both datasets.

For the Acupuncture dataset, where approximately 25% of cases have
missing data, we increased K from the default 5 to 20 and 25. For the
VITAL dataset, which has around 50% missingness, we increased K from 5
to 20 and then to 50.

### Sensitive analysis - Amy

As mentioned in our introduction, missing data mechanisms are
unverifiable. While it is most common to assume that data are MAR, in
reality, the mechanism could fall under MNAR. We conducted a sensitivity
analysis to assess the robustness of the estimated treatment effect
under alternative missingness assumptions. Sensitivity analysis is
recommended by the National Research Council (2010) as a best practice
for handling missing data; however, it is rarely conducted in practice.
This may reflect an underappreciation of the impact of missing data in
many studies, as well as the complexity and specialised expertise
required to perform such analyses.

A systematic review by Fiero et al. (2016) examined 86 cluster
randomised trials published between 2013 and 2014. Among the 80 trials
that reported missing outcome data, only 14 (18%) reported conducting a
sensitivity analysis.

In our study, we implemented multiple imputation with $\delta$
-adjustment as our sensitivity analysis approach. In this method, a
constant shift $\delta$ is added to the imputed values after the initial
imputation step. This enables us to explore a range of plausible
scenarios by re-estimating the treatment effect under different
assumptions about the missing data mechanism. By applying these shift
parameters, we can model situations in which participants with missing
data are assumed to have systematically better or worse outcomes than
those observed, thereby testing the stability of our conclusions when
relaxing the MAR assumption.

The shift can be broken down in to:

$$\delta = \delta_\mathrm{MAR} + \delta_\mathrm{MNAR}$$ where; -
$\delta_\mathrm{MAR}$: mean difference caused by the predictors in the
imputation models - $\delta_\mathrm{MNAR}$: mean difference caused by an
additional non-ignorable part of the imputation model

Essentially, $\delta$ is the mean difference between the imputed values
and the true values that we did not observe.

Deciding which $\delta$ values are suitable is subjective. We need to
include $\delta=0$ as that assumes the data is MAR which acts as our
baseline, and we add the non-zero $\delta$ values to test the
sensitivity to the MNAR assumption. We decided to base our $\delta$
values on the standard errors of the treatment effect estimates, for
which ranges between 1.08 and 1.27 in the acupuncture trial data and
between 0.76 and 1.21 in the VITAL data.

We decided an increasing and decreasing range that do not deter much
from these values should suffice. Our chosen $\delta$ values are;

$\delta_\mathrm{Acupuncture} = -2.5,-2,-1.5,-1,-0.5,0,0.5,1,1.5,2,2.5$
$\delta_\mathrm{VITAL} -2.5,-2,-1.5,-1,-0.5,0,0.5,1,1.5,2,2.5$

**Sensitivity Analysis; the process**

The mice() function was used on the wide format of the acupuncture trial
data with predictors `group`, `pk1`. An initial iteration was conducted
with m=1 in order to extract the post processing command. This allows us
to add the $\delta$ to the missing values after they have been imputed.
This is looped over the range of $\delta$ values.

Next we performed the main imputation, indicating post processing, using
predictive mean matching method and maxit=10 to indicate maximum of 10
iterations. After imputation, a regression model like our substantive
model is fitted, $\beta_0 + \beta_1group+\beta_2pk1_i+\varepsilon_i$,
and the estimates are pooled together.

\newpage

```{r eval=FALSE, echo=TRUE}
#predictors
inlist <- c("group", "pk5", "pk1")
#predictor matrix
pred_cat <- quickpred(acu_wide, minpuc = 0.5, include = inlist)
#initial imputation to extract post processing 
imp.default_cat <- mice(acu_wide, m = 1, maxit = 1, predictorMatrix = pred_cat, seed = 123, print= FALSE)
post_cat <- imp.default_cat$post
imp.all.undamped_cat <- vector("list", length(delta_acu))

#loop over delta values
for (i in 1:length(delta)) {
  d <- delta[i]
  cmd <- paste0(
    "idx <- which(data[,'group'] == 1 & is.na(data[,'pk5'])); ",
    "imp[[j]]$pk5[idx] <- imp[[j]]$pk5[idx] + ",d, ";") #treatment group and imputing pk5, adding delta d
  post_cat["pk5"] <- cmd
  imp_cat <- mice(acu_wide, pred = pred_cat, post = post_cat, maxit = 10,
                  seed = i * 22, print=FALSE)
  imp.all.undamped_cat[[i]] <- imp_cat
}
delta_results_cat <- data.frame()

for (i in seq_along(imp.all.undamped_cat)) {
  imp_cat <- imp.all.undamped_cat[[i]]
  d <- delta[i]
  fit_cat <- with(imp_cat, lm(pk5 ~ group + pk1)) #imputation model
  pooled_cat <- pool(fit_cat) #pooling estimates
  est_cat <- tidy(pooled_cat, conf.int = TRUE) %>%
    filter(term == "group") %>% 
    select(estimate, std.error, conf.low, conf.high, p.value) %>%
    mutate(delta_acu = d)
    
  delta_results_cat <- bind_rows(delta_results_cat, est_cat)
}
```

We repeated this process exactly on the `acu_wide` data in which we
added a filter(group==0) in order to extract the results in the control
group for comparison, creating `acu_wide_placebo`.

A similar process was implemented on to the VITAL trial data in wide
format. We chose our inlist predictors `pain_base`,
`vitdactive`,`fishoilactive`, and our outcome as `pain_yr4`. After
pooling our estimates, we filtered out the treatment groups by setting
'term == fishoilactive' and 'term == vitdactive' so we can observe the
results by treatment. For the control group we filtered the `vital_wide`
data as fishoilactive==0 and vitdactive==0.

We performed sensitivity analysis in a similar fashion to test the
robustness of the treatment effect on our alternative estimand. This was
more intensive as we had to implement the multiple imputation on the
wide format then transform the data in to long format.

For the acupuncture trial; we used our to_long_format_acu_cont_MICE()
function post imputation to transform to long format. Due to the nature
of the long format, we used the linear mixed effects model to estimate
our treatment effect before pooling the estimates. Our imputation model
here is;

$\text{pk\_score}_{ij} = \beta_0 + \beta_1 \text{group} + \beta_2  \text{time\_c} + \beta_3  (\text{group}  \text{time\_c}) + \beta_4  \text{pk1} + b_0 + \varepsilon_{ij}$

\newpage

For the VITAL data; we created a similar function to our
to_long_format_vital_MICE() function with a minor adjustment;

```{r eval=FALSE, echo=TRUE}
to_long_format_vital_mice_SA <- function(data_wide) {
  data_wide %>%
    pivot_longer(cols = matches("_yr[[:digit:]]$"),
                 names_to = c(".value", "time"), 
                 names_sep = "_") %>%
    group_by(.imp) %>%
    mutate(.id = 1:n()) %>%
    mutate(time_contin = as.integer(gsub("yr", "", time)),
           time_contin_cent = time_contin - 4)
}
```

We used
$pain_{ij}=\beta_0+\beta_1fishoil_i+\beta_2time_{ij}+\beta_3(fishoil_itime_{ij})+\beta_4vitd_i+\beta_5vitd_itime_{ij}+\beta_6(pain_base)_i+b_{0i}+\varepsilon_{ij}$
as our imputation model in each of the imputed data sets.

```{r eval=FALSE, echo=TRUE}
#predictors
inlist <- c("fishoilactive", "vitdactive", "pain", "pain_base", "time_contin", "Subject_ID") 
pred_cont <- quickpred(vital_wide, minpuc = 0.5, include = inlist) #prediction matrix
imp.all.undamped_cont <- vector("list", length(delta_vital))
delta_results_cont_fishoil <- data.frame()

for (i in seq_along(delta_vital)) {
  d <- delta_vital[i]
  imp_init <- mice(vital_wide, m = 5, maxit = 1, predictorMatrix = pred_cont, seed = 100 + i, print = FALSE)
  post_cont <- imp_init$post
  method_cont <- imp_init$method
  years <- paste0("pain_yr", 1:4) #imputing over 4 years
  method_cont[years] <- "pmm" # predictive mean matching
  method_cont["pain_base"] <- "pmm" #imputing base pain
  for (yr in years) {
    post_cont[yr] <- paste0(
      "idx <- which(data[,'fishoilactive'] == 1 & data[,'vitdactive'] == 0 & is.na(data[,'", yr, "'])); ",
      "imp[[j]]$", yr, "[idx] <- imp[[j]]$", yr, "[idx] + ", d, ";" #fishoil only 
    )
  }
  imp_wide <- mice(vital_wide, m = 5, maxit = 10, predictorMatrix = pred_cont,
                   post = post_cont, method=method_cont, seed = 200 + i, print = FALSE)
  imp.all.undamped_cont[[i]] <- imp_wide 
  imp_data_long <- complete(imp_wide, action = "long", include = TRUE) %>%
    to_long_format_vital_mice_SA() #transform to long data
  imp_data_list <- split(imp_data_long, imp_data_long$.imp)
  
  fit_list <- lapply(imp_data_list, function(df) {
    lmer(pain ~ fishoilactive*time_contin + vitdactive*time_contin + pain_base + 
           (1 | Subject_ID), data = df)
  })
  pooled_cont <- pool(fit_list)
  est_cont <- tidy(pooled_cont, conf.int = TRUE) %>%
    filter(term == "fishoilactive") %>%
    select(estimate, std.error, conf.low, conf.high) %>%
    mutate(delta_vital = d)
  
  delta_results_cont_fishoil <- bind_rows(delta_results_cont_fishoil, est_cont)
}
```

# Result

## Comparing different methods

We begin by comparing our anchoring methods with more statistically
principled approaches.

```{r fig.width=5, fig.height=3, fig.cap="Methods from top to bottom: CCA = complete case analysis; LOCR = last observation carry forward; MO = mean observation; MI = multiple imputation with linear regression; LME = linear mixed-effects model without imputation; MI+LME = multiple imputation followed by linear mixed-effects model."}
print(acu_plot_categorical) 
print(vital_plot_categorical)
```

## Changing estimand

In this section, we explore how changing the estimand (how time is
modeled) to see how that affects our conclusions.

```{r fig.width=5, fig.height=3, fig.cap="LME: linear mix model without imputations; MI+LME: multiple imputation and linear miex model; Purple line (Original Estimand) estimand: The mean difference of pain score at the end of study conditional on baseline pain score, treating time as categorical variable; Green line (Changed Estimand) estimand: The mean difference of pain score at the end of study conditional on baseline pain score, treating time as continuous variable."}
print(acu_plot_compare)
print(vital_plot_compare)
```

## Changing imputation methods and imputation numbers

In this session. We keep our estimand treating time as a categorical
variable and using multiple linear regression as substantive model.

We first compared different imputation methods within the FCS framework,
using 5 imputations.

```{r fig.width=5, fig.height=3}
print(acu_plot_imp) 
print(vital_plot_impt)
```

Next, we assess the impact of changing the number of imputations, using predictive mean method. As introduced in the previous session, we used $K=5$, $K=20$, and $K=100*\lambda$ (Which is 25 and 50 for Acupuncture and VITAL data sets accordingly).

```{r fig.width=5, fig.height=3}
print(acu_plot_imp_k)
print(vital_plot_impt_k)
```

## Sensitive analysis

```{r fig.width=5, fig.height=4}
SA_acu_all
SA_vital_all_plot
```

The forest plot aims to illustrate the robustness of the estimated
treatment effects under varying missing data assumptions using $\delta$
adjustment sensitivity analysis. In figure x which shows the acupuncture
trial results, we can observe the conditional expectation of headache
severity at the final time point of 12 months given the baseline
headache severity and acupuncture treatment under different $\delta$
shifts. Similarly we can also observe the conditional expectation of
headache severity at the final time point in the control group, given
baseline headache severity only and no treatment.

Figure Y shows the VITAL results after applying the sensitivity
analysis. In the fish oil panel, we observe the conditional expectation
of knee pain score at 4 years given baseline knee pain score and fish
oil treatment under the different $\delta$ shifts. The vitamin D panel
shows similar conditional expectation, but the control group shows the
conditional expectation of knee pain score given baseline knee pain
score and no treatment.

The error bars represent the 95% confidence intervals.

The $\delta$ parameter shifts the imputed values to simulate scenarios
where the missing data are systematically better or worse than the
observed ones. Positive $\delta$ values imply better unobserved outcomes
while negative values imply worse. The consistency of estimates across
the $\delta$ values suggests the treatment effect is robust to
departures from the MAR assumption.

\newpage

```{r fig.width=4, fig.height=2}
cat_heatmap_acu_plot
cont_heatmap_acu_plot
cat_heatmap_fishoil_plot
cont_heatmap_fishoil_plot
cat_heatmap_vitd_plot
cont_heatmap_vitd_plot
```

We created heatmaps plotting the $\delta$ values for both the control
group and treatment group to visualise the change in treatment effect
when varying the $\delta$ adjustments under MNAR. The aim here is to
show the robustness of the initial estimated treatment effect when
assumed under a different mechanism. A similar range of colours would be
interpretable as a strong estimate as there was not much change in the
estimate after adding the $\delta$ shifts.

# Discussion

Across all analyses, no meaningful difference was observed in the final
treatment effect estimates across the various methods and settings
explored.

However, several points are worth noting:

-   In Figures 1 and 2, we observe that methods involving multiple
    imputation (MI) show more variation in the estimated treatment
    effects. This is expected, as MI leverages all available observed
    data, while the other four methods (CAA, LOCF, MO, and LME without
    imputation) use only a subset—typically including only group and
    baseline pain score.

-   When we changed the estimand to treat time as a continuous variable,
    the impact was minimal in the Acupuncture study, which is
    unsurprising given that it includes only two follow-up time points.
    In such cases, modeling time continuously offers limited additional
    information.

It is important to emphasize that the goal of this project is not to
revise prior conclusions from the original studies, but rather to gain
practical insights by applying and comparing different methods for
handling missing data.

The fact that our results are broadly consistent across methods should
not be taken to imply that simpler approaches like CCA are equivalent to more principled methods. Even in a
similar future study, the same results may not hold, especially under
different missing data patterns or assumptions.

This project has highlighted that there are many valid approaches to
handling missing data, and that these can be combined in different ways
depending on the context. It is crucial to understand which methods are
most appropriate under which conditions.

For example:

-   LME is a powerful tool for analyzing longitudinal data and can yield
    unbiased results even without imputation—if the assumptions are met.
    However, in the Acupuncture dataset, its utility is limited due to
    the small number of timepoints and its inability to incorporate
    other post-randomisation variables like pain frequency.

-   In the VITAL dataset, although there are more timepoints, baseline pain scores are sometimes missing. As a result, when fitting LME without imputation, individuals with missing baseline values are excluded. This exclusion becomes more severe as more covariates with missing data are included, potentially leading to substantial loss of information. While it is theoretically possible to incorporate post-randomisation variables into an LME model by extending the covariance structure, this approach is complex and less flexible. In contrast, using MI provides a more straightforward way to incorporate auxiliary post-randomisation information during the imputation step, without requiring modifications to the model structure.

## Limitations of the project

There are several limitations stemming from the inherent characteristics
of the datasets used in this project.These limitations are not a result
of analytic choices but rather reflect the practical boundaries of the
available data.

-   *Limited follow-up in Acupuncture dataset:* The Acupuncture study
    includes only two follow-up timepoints (3 and 12 months), which
    limits the ability to model time flexibly or detect subtle
    longitudinal trends. As a result, methods like linear mixed-effects
    models offer limited additional benefit in this context.

-   *Weak treatment effect in VITAL:* The VITAL dataset shows relatively
    small therapeutic effects, making it difficult to detect meaningful
    differences between methods. This also reduces the practical impact
    of using more advanced imputation strategies.

-   *No known data-generating mechanism:* Since we are working with one-off samples from an unknown data-generating process and a true, unobserved treatment effect—rather than simulated data—we cannot evaluate the performance of different analytical methods against a known ground truth. This limits our ability to draw general conclusions about their long-run properties.

## Future work

In addition, there are methodological aspects that could be improved or
explored further in future work. These include choices related to model
specification, variable adjustment, and the diversity of imputation
techniques employed. Addressing these areas could enhance the robustness
and generalizability of findings, especially in more complex or
data-rich clinical contexts.

-   *Model specification:* Both linear regression and LME rely on assumptions such as linearity and normally distributed residuals. While these assumptions were not formally tested, the sample size in the VITAL dataset is likely large enough for the central limit theorem to provide valid inference. That said, future work could benefit from visual checks of residuals and distributions of pain scores at each timepoint to give a clearer sense of model fit and outcome behaviour.

-   *Imputation strategy:* In the most of our analysis, we used predictive mean matching for imputing missing variables. While this is a valid and robust approach, many other imputation models as we introduced were not explored in detail. These could be considered in future work to assess sensitivity to imputation method.

-   *Covariate adjustment in substantive models:* Our analysis models adjusted only for treatment group and baseline pain score in our substantive model to ensure comparability across methods. This low-level of adjustment in the substantive model is the reason we observed larger discrepancies between results using MI and those not using MI, as a fuller set of covariates was used during MI. Future work could explore how more complex substantive model specifications interacts with MI and influences treatment effect estimates.