---
title: "missingness analysis"
author: "Amy, Kent"
date: "2025-02-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(mice)
library(tidyverse)
library(haven)
library(lme4)
```

## Acupuncture

```{r}
# Load data
acu_wide <- read_csv("Data/acupuncture.csv",show_col_types = FALSE)
head(acu_wide)

# Long format
acu_long <- acu_wide %>%
  pivot_longer(
    cols = c('pk2', 'pk5'), 
    names_to = c('pk_time'),
    values_to = "pk_score"
  ) %>%
  pivot_longer(
    cols = c('f2', 'f5'), 
    names_to = c('freq_time'),
    values_to = "freq_score",
    names_repair = "unique"
  ) %>%
  filter((pk_time == "pk2" & freq_time == "f2") | (pk_time == "pk5" & freq_time == "f5")) %>%
  mutate(
    time = case_when(
      pk_time == "pk2" ~ "3m",
      pk_time == "pk5" ~ "12m"
    )
  ) %>%
  select(-c(pk_time, freq_time))

head(acu_long)




# LOCF function
LOCF <- function(data, columns){
  for (col in columns) {
    for (i in 1:nrow(data)) {  
      if (is.na(data[i, col])) {
        data[i, col] <- data[i, col - 1]  
      }
    }
  }
  return(data)  
}

acu_filtered<- acu_long%>%
  filter(time != "3m")
acu_LOCF <- LOCF(data = acu_filtered, columns = c(10,11))
head(acu_LOCF)

# MEAN IMPUTATION FUNCTION 
mean_impute <- function(data) {
  data <- data.frame(
    lapply(data, function(x) {
      if (is.numeric(x)) {
        return(ifelse(is.na(x), mean(x, na.rm = TRUE), x))
      } else {
        return(x)
      }
    })
  )
  return(data)
}
acu_simp<-mean_impute(acu_wide)
head(acu_simp)
```

```{r warning=FALSE}
acu_filtered<- acu_long%>%
  filter(time != "3m")

# CAA with filtered data 
acu_complete <- lm(pk_score ~ group + pk1 , data=acu_filtered)
tidy(acu_complete)

# LOCF with long filtered data
LOCF_acu_model<- lm(pk_score ~ group + pk1, data=acu_LOCF) #baseline only #wide data
tidy(LOCF_acu_model)

# MM with wide data to be transformed 
acu_mice <- mice(acu_wide, m = 5, method = 'pmm', seed = 123)
acu_mice_model <- with(acu_mice, lm(pk5 ~ group + pk1)) #baseline only - to be used in linear mixed effects model 
acu_pooled_model<- pool(acu_mice_model)
tidy(acu_pooled_model)

# LME with long data 
acu_LME<- lmer(pk_score ~ group + time + (1| id), data = acu_long)
summary(acu_LME)
```

|Method|coefficient|SE|p|
|--|--|--|--|
|CAA|-4.5868409	|1.25177152|0.00029|
|LOCF|-3.3682609	|1.02885179	|0.00115|
|MM|-2.7799226|1.27952587|0.0305|
|LMM|-5.63|1.67|N/A|

## Load data



## VITAL DATA ANALYSIS

```{r}
#Vital
data_vitalkp <- read_sas("Data/vitalkneepain111623.sas7bdat", NULL)
data_main <- read_sas("Data/VITAL_trial_NEJM_2022.sas7bdat", NULL)
vital_wide<- data_main %>%
  dplyr::select(Subject_ID, vitdactive, fishoilactive, sex, ageyr, ) %>%
  right_join(data_vitalkp, by = "Subject_ID")
```

## VITAL DATA TO LONG FORMAT

```{r message=FALSE}
# Vital
vital_long<-vital_wide%>%
    pivot_longer(
    cols = c('pain_yr1','pain_yr2', 'pain_yr3', 'pain_yr4'), 
    names_to = c('pain_time'),
    values_to = "pain_score"
  ) %>%
    pivot_longer(
    cols = c('stiffness_yr1', 'stiffness_yr2', 'stiffness_yr3', 'stiffness_yr4'), 
    names_to = c('stiffness_time'),
    values_to="stiffness_score",
    names_repair = "unique"
    ) %>%
    pivot_longer(
    cols = c('function_yr1', 'function_yr2', 'function_yr3', 'function_yr4'), 
    names_to = c('function_time'),
    values_to = "function_score",
    names_repair = "unique"
    ) %>%
  filter((pain_time == "pain_yr1" & stiffness_time == "stiffness_yr1" & function_time == "function_yr1")| (pain_time == "pain_yr2" & stiffness_time == "stiffness_yr2" & function_time == "function_yr2") | (pain_time == "pain_yr3" & stiffness_time == "stiffness_yr3" & function_time == "function_yr3") | (pain_time == "pain_yr4" & stiffness_time == "stiffness_yr4" & function_time == "function_yr4")) %>%
  mutate( 
    time=case_when(
      pain_time == "pain_yr1" ~ "Year1",
       pain_time == "pain_yr2" ~ "Year2",
       pain_time == "pain_yr3" ~ "Year3",
        pain_time == "pain_yr4" ~ "Year4",
      stiffness_time == "stiffness_yr1" ~ "Year1",
       stiffness_time == "stiffness_yr2" ~ "Year2",
        stiffness_time == "stiffness_yr3" ~ "Year3",
         stiffness_time == "stiffness_yr4" ~ "Year4",
      function_time == "function_yr1" ~ "Year1",
      function_time == "function_yr2" ~ "Year2",
      function_time == "function_yr3" ~ "Year3",
      function_time == "function_yr4" ~ "Year4"
    )
  ) %>%  select(-c(pain_time, stiffness_time, function_time))
  
head(vital_long)
```

## Complete Case Analysis - just use linear model as it removes rows with NA values 

```{r}
#CCA
vitd_complete <- lm(pain_yr4 ~ vitdactive*pain_yr1, data=vital_wide) 
fishoil_complete <- lm(pain_yr4 ~ fishoilactive*pain_yr1, data=vital_wide) 
tidy(vitd_complete)
tidy(fishoil_complete)
```

## Last Observation Carried Forward

```{r}
#LOCF function
LOCF <- function(data, columns){
  for (col in columns) {
    for (i in 1:nrow(data)) {  
      if (is.na(data[i, col])) {
        data[i, col] <- data[i, col - 1]  
      }
    }
  }
  return(data)  
}
```

```{r}
#Vital
vital_LOCF<- vital_wide
vital_LOCF<- LOCF(data = vital_LOCF, columns = c(9, 10,11,15,16,17,18,19,20,22,23))
head(vital_LOCF)
```

## SIMPLE MEAN IMPUTATION
```{r}
mean_impute <- function(data) {
  data <- data.frame(
    lapply(data, function(x) {
      if (is.numeric(x)) {
        return(ifelse(is.na(x), mean(x, na.rm = TRUE), x))
      } else {
        return(x)
      }
    })
  )
  return(data)
}
vital_simp<-mean_impute(vital_wide)
head(vital_simp)
```


## MICE package with vital data

```{r warning}
#multiple imputation 
vital_mice <- mice(vital_wide, m = 5, method = 'pmm', seed = 123)
vital_mice_data <- complete(vital_mice, action = "long", include = TRUE)
vital_mice_data
```


```{r warning}
#vitaminD lm
vitD_mice_model <- with(vital_mice, lm(pain_yr4 ~ vitdactive * pain_base)) #baseline only - to be used in linear mixed effects model 
vitD_pooled_model<- pool(vitD_mice_model)
tidy(vitD_pooled_model)

#fishoil lm 
fishoil_mice_model <- with(vital_mice, lm(pain_yr4 ~ fishoilactive * pain_base)) #baseline only - to be used in linear mixed effects model 
fishoil_pooled_model<- pool(fishoil_mice_model)
tidy(fishoil_pooled_model)

```


## Mixed Effects Model

```{r}
lmer(pain_score~vitdactive*time + (1|Subject_ID), data=vital_long)
fishoil_LME<-lmer(pain_score~fishoilactive*time+ (1|Subject_ID), data=vital_long)
summary(vitD_LME)
summary(fishoil_LME)
```


## Mixed Effects models using MICE 

```{r}

#vitaminD lmer
#vitD_mice_lme <- with(vital_mice, lmer(pain_yr4 ~ vitdactive * pain_base + (1|Subject_ID))) #baseline only - to be used in linear mixed effects model 
#vitD_pooled_lme<- pool(vitD_mice_lme)
#tidy(vitD_pooled_lme)

#fishoil lmer
#fishoil_mice_lme <- with(vital_mice, lmer(pain_yr4 ~ fishoilactive*pain_base + (1|Subject_ID))) #baseline only - to be used in linear mixed effects model 
#fishoil_pooled_lme<- pool(fishoil_mice_lme)
#tidy(fishoil_pooled_lme)
```


#EXTRA CODE

```{r warning=FALSE
head(acupuncture)
acu_long <- reshape(acupuncture, 
                    varying = c('pk1', 'pk2', 'pk5', 'f1', 'f2', 'f5'),
                    v.names = 'value', 
                    times = c('pk1', 'pk2', 'pk5', 'f1', 'f2', 'f5'),
                    direction = 'long'

vital_long<-reshape(data_vital, 
                    varying = c('vitdactive', 'fishoilactive', 'sex', 'ageyr', 'pain_base', 
                                'stiffness_base', 'function_base', 'pain_yr1', 'stiffness_yr1', 'function_yr1', 'pain_yr2', 'stiffness_yr2', 'function_yr2','pain_yr3', 'stiffness_yr3', 'function_yr3', 'pain_yr4', 'stiffness_yr4', 'function_yr4', 'KneesReplacedV2','unikneepain','bikneepain','kneepainfreq','tylenolbase','nsaidsbase','strongerbase','tkrf'),
                    v.names = 'value', 
                    times = c('vitdactive', 'fishoilactive', 'sex', 'ageyr', 'pain_base', 
                                'stiffness_base', 'function_base', 'pain_yr1', 'stiffness_yr1', 'function_yr1', 'pain_yr2', 'stiffness_yr2', 'function_yr2','pain_yr3', 'stiffness_yr3', 'function_yr3', 'pain_yr4', 'stiffness_yr4', 'function_yr4', 'KneesReplacedV2','unikneepain','bikneepain','kneepainfreq','tylenolbase','nsaidsbase','strongerbase','tkrf'),
                    direction = 'long')
acu_long

functions
```{r
LOCF <- function(x) {
  v <- !is.na(x)
  c(NA, x[v])[cumsum(v)+1]
}
```

```{r
LOCF <- function(x) {
    LOCF <- max(which(!is.na(x)))
    x[LOCF:length(x)] <- x[LOCF]
    return(x)
}
data.frame(lapply(acupuncture, LOCF))
```

```{r
#locf function on vital data zoo package
LOCF_vital<-as.data.frame(na.locf(vital_wide, fromLast=TRUE))
```

```{r
#locf function on acupuncture data zoo package
LOCF_acupuncture<-as.data.frame(na.locf(acu_wide, fromLast=TRUE))
```

```{r
#models from zoo package locf
LOCF_fishoil_model<- lm(pain_yr4~fishoilactive+sex+ageyr+pain_base, data=LOCF_vital)
LOCF_vitD_model<-lm(pain_yr4~vitdactive+sex+ageyr+pain_base, data=LOCF_vital)
```

```{r
#table of complete case, locf and regular model for acupuncture 
tab_model(acu_complete, acu_LOCF )
```
