---
title: "missingness analysis"
author: "Amy, Kent"
date: "2025-02-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(mice)
library(tidyverse)
library(haven)
library(lme4)
library(broom.mixed)
source("R scripts/Functions.R")
source("R scripts/VITAL data formats.R")
source("R scripts/Acupuncture data formats.R")
```

To do
- Apply Neil's
- Workout original MM method
- Reformat Table
  - Estimate
    - Linear Model
    - LME (Time No/Category/Continue)
  - Missing Method
    - Do Nothing
    - Simple imputation
    - MI
  - Imputation Method
    - TBD
- Get forest graph

## Acupuncture

```{r warning=FALSE}

# CAA
acu_CAA <- lm(pk5 ~ group + pk1 , data = acu_wide)
tidy(acu_CAA)
acu_CAA_CI <- confint(acu_CAA)
acu_CAA_CI

# LOCF
acu_LOCF_data <- LOCF(data = acu_wide, columns = c(10,11))
acu_LOCF <- lm(pk5 ~ group + pk1, data = acu_LOCF_data)
tidy(acu_LOCF)
acu_LOCF_CI <- confint(acu_LOCF)
acu_LOCF_CI

# Simple mean imputation
acu_SPM_data <- mean_impute(acu_wide)
acu_SPM <- lm(pk5 ~ group + pk1, data = acu_SPM_data)
tidy(acu_SPM)
acu_SPM_CI <- confint(acu_SPM)
acu_SPM_CI

# MI with mice default method
acu_MICE_default <- with(acu_mice, lm(pk5 ~ group + pk1)) 
acu_MICE_default_pool <- pool(acu_MICE_default)
tidy(acu_MICE_default_pool) #CI?

# LME with no imputation
acu_LME <- lmer(pk_score ~ group + time + (1| id), data = acu_long) # it is right to + time?
acu_LME_long_CI <- confint.merMod(acu_LME)
tidy(acu_LME)
acu_LME_long_CI
```

**updated table**

|Method|coefficient|SE|p|
|--|--|--|--|
|CAA|-4.59|1.25|0.000294|
|LOCF|-3.37|1.03|0.00116|
|MI|-2.70|1.26|0.0329|
|LMM|-5.63|1.67|N/A|

#MICE DATA PROCESSING ACUPUNCTURE

```{r}
acu_lmem_fit1 <- with(acu_mi_obj_long, lmer(pk_score ~ group*pk1 + (1|id)))
acu_lmem_fit_pool1 <- pool(acu_lmem_fit1)
acu_lmem_summary1 <- summary(acu_lmem_fit_pool1, conf.int = TRUE, conf.method = "Wald")
acu_lmem_summary1 %>% filter(term %in% "group")


acu_lmem_fit2 <- with(acu_mi_obj_long, lmer(pk_score ~ group+pk1 + (1|id)))
acu_lmem_fit_pool2 <- pool(acu_lmem_fit2)
acu_lmem_summary2 <- summary(acu_lmem_fit_pool2, conf.int = TRUE, conf.method = "Wald")
acu_lmem_summary2 %>% filter(term %in% "group")

acu_lmem_fit3 <- with(acu_mi_obj_long, lmer(pk_score ~ group+freq_score*pk1 + (1|id)))
acu_lmem_fit_pool3 <- pool(acu_lmem_fit3)
acu_lmem_summary3 <- summary(acu_lmem_fit_pool3, conf.int = TRUE, conf.method = "Wald")
acu_lmem_summary3 %>% filter(term %in% "group")

acu_lmem_fit4 <- with(acu_mi_obj_long, lmer(pk_score ~ group+freq_score*pk1+migraine + (1|id)))
acu_lmem_fit_pool4 <- pool(acu_lmem_fit4)
acu_lmem_summary4 <- summary(acu_lmem_fit_pool4, conf.int = TRUE, conf.method = "Wald")
acu_lmem_summary4 %>% filter(term %in% "group")

acu_lmem_fit5 <- with(acu_mi_obj_long, lmer(pk_score ~ group+freq_score*pk1+pk1*chronicity + (1|id)))
acu_lmem_fit_pool5 <- pool(acu_lmem_fit5)
acu_lmem_summary5 <- summary(acu_lmem_fit_pool5, conf.int = TRUE, conf.method = "Wald")
acu_lmem_summary5 %>% filter(term %in% "group")
```
#Non imputed

```{r warning=FALSE}
acu_lmem_fit1 <- with(acu_long, lmer(pk_score ~ group*pk1 + (1|id)))
acu_lmem_summary1 <- summary(acu_lmem_fit1, conf.int = TRUE, conf.method = "Wald")
confint(acu_lmem_fit1)

acu_lmem_fit2 <- with(acu_long, lmer(pk_score ~ group+pk1 + (1|id)))
acu_lmem_summary2 <- summary(acu_lmem_fit2, conf.int = TRUE, conf.method = "Wald")
acu_lmem_summary2$coefficients["group", ]
confint(acu_lmem_fit2)

acu_lmem_fit3 <- with(acu_long, lmer(pk_score ~ group+freq_score*pk1 + (1|id)))
acu_lmem_summary3 <- summary(acu_lmem_fit3, conf.int = TRUE, conf.method = "Wald")
acu_lmem_summary3$coefficients["group", ]
confint(acu_lmem_fit3)

acu_lmem_fit4 <- with(acu_long, lmer(pk_score ~ group+freq_score*pk1+migraine + (1|id)))
acu_lmem_summary4 <- summary(acu_lmem_fit4, conf.int = TRUE, conf.method = "Wald")
acu_lmem_summary4$coefficients["group", ] 
confint(acu_lmem_fit4)

acu_lmem_fit5 <- with(acu_long, lmer(pk_score ~ group+freq_score*pk1+pk1*chronicity + (1|id)))
acu_lmem_summary5 <- summary(acu_lmem_fit4, conf.int = TRUE, conf.method = "Wald")
acu_lmem_summary5$coefficients["group", ] 
confint(acu_lmem_fit5)
```


## VITAL DATA ANALYSIS


```{r}
#CCA
vital_complete <- lm(pain_yr4 ~ fishoilactive + vitdactive + pain_base, data = vital_wide) 
vital_complete_summary <- tidy(vital_complete, conf.int = TRUE, conf.method = "Wald")

#MOCF
vital_MOCF <- mean_impute(vital_wide)
vital_MOCF <- lm(pain_yr4 ~ fishoilactive + vitdactive + pain_base, data = vital_MOCF)
vital_MOCF_summary <- tidy(vital_MOCF, conf.int = TRUE, conf.method = "Wald")

#MI+SLR
vital_MI_SLR <- with(vital_mice, lm(pain_yr4 ~ fishoilactive + vitdactive  + pain_base ))
vital_MI_SLR_pool <- pool(vital_MI_SLR)
vital_MI_SLR_summary <- summary(vital_MI_SLR_pool, conf.int = TRUE, conf.method = "Wald")

#LME
vital_LME <- lmer(pain ~ fishoilactive * time_contin_cent + vitdactive * time_contin_cent + pain_base + 
                       (time_contin_cent|Subject_ID), 
                     data = vital_long)
vital_LME_summary <- tidy(vital_LME, conf.int = TRUE, conf.method = "Wald")

#MI+LME
fishoil_MI_LME <- with(vital_mice_obj_long, 
                     lmer(pain ~ fishoilactive * time_contin_cent + vitdactive * time_contin_cent + pain_base + 
                            (time_contin_cent|Subject_ID)))
vital_MI_LME_pool <- pool(fishoil_MI_LME)
vital_MI_LME_summary <- summary(vital_MI_LME_pool, conf.int = TRUE, conf.method = "Wald")

vital_complete_summary %>% filter(term %in% c("fishoilactive", "vitdactive"))
vital_MI_LME_summary %>% filter(term %in% c("fishoilactive", "vitdactive"))
vital_MI_SLR_summary %>% filter(term %in% c("fishoilactive", "vitdactive"))
vital_LME_summary %>% filter(term %in% c("fishoilactive", "vitdactive"))
vital_MI_LME_summary %>% filter(term %in% c("fishoilactive", "vitdactive"))
```

## Mixed Effects models using MICE 

```{r}

#vitaminD lmer
#vitD_mice_lme <- with(vital_mice, lmer(pain_yr4 ~ vitdactive * pain_base + (1|Subject_ID))) #baseline only - to be used in linear mixed effects model 
#vitD_pooled_lme<- pool(vitD_mice_lme)
#tidy(vitD_pooled_lme)

#fishoil lmer
#fishoil_mice_lme <- with(vital_mice, lmer(pain_yr4 ~ fishoilactive*pain_base + (1|Subject_ID))) #baseline only - to be used in linear mixed effects model 
#fishoil_pooled_lme<- pool(fishoil_mice_lme)
#tidy(fishoil_pooled_lme)
```


## FOREST PLOT


```{r}
library(meta)

acu_trtmt_data <- tibble(treatment  = c(-4.59, -3.37, -3.37, -5.63),
                            lower = c(-7.05, -5.39, -5.79, -8.9),
                            upper = c(-2.123, -1.35, -1.53, -2.35),
                            method = c("Linear (No Imputation)", "Linear (Last Observation Carried Forward)", "Linear (Simple Mean Imputed)", "Linear Mixed Effect (No Imputation)"))

ggplot(acu_trtmt_data, aes(x = treatment, y = method, xmin = lower, xmax = upper)) +
  geom_point(size = 4, color = "deeppink") + geom_vline(xintercept = 0, linetype="dashed", color="red") +
  geom_errorbarh(aes(xmin = lower, xmax = upper), height = 0.4, color = "black") + 
  labs(
    x = "Treatment Effect",
    y = "Method",
    title = "Treatment effect with CI") +
  theme_minimal() 
  
```

```{r}
ACU_MICE_DATA <- tibble(estimate  = c(1.0071, -2.246864, -1.944244, -1.929348, -1.938256, -0.5887, -4.4087, -2.051, -2.049, -2.049),
                            lower = c(-3.13397, -4.7174, -3.12579, -3.109794, -3.119116, -3.294, -6.445, -3.4454, -3.4442, -3.4572),
                            upper = c(5.148172, 0.3763, -0.7626976, -0.7489025, -0.7573948, 4.471, -2.373, -0.656, -0.6556, -0.6782),
                            model = c("1","2","3","4","5","1","2","3","4","5"),
                            impute = c("imputed", "imputed", "imputed", "imputed", "imputed", "non-imputed","non-imputed","non-imputed","non-imputed","non-imputed"))

ggplot(ACU_MICE_DATA, aes(x = estimate, y = model, xmin = lower, xmax = upper, colour=impute)) +
  geom_point(size = 3) + geom_vline(xintercept = 0, linetype="dashed", color="red") +
  geom_errorbarh(aes(xmin = lower, xmax = upper), height = 0.4) + scale_color_manual(name="MICE",values = c("imputed" = "lawngreen", "non-imputed" = "olivedrab"), labels=c("5 Imputed Datasets","Observed Data")) +
  labs(
    x = "Treatment Effect",
    y = "Model",
    title = "Estimated Treatment Effects with Wald 95% Confidence Intervals") + theme_minimal()
#+ theme(
   # plot.background = element_rect(fill = "#a80050", color = "transparent"),  
   # panel.background = element_rect(fill = "white", color = "transparent"),
   # legend.title = element_text(size = 12, face = "bold"),
   # legend.background = element_rect(fill = "white", color = "transparent")  
 # ) 
```

