---
title: Estimating Treatment Effects in Clinical Trials with Missing Data
author:
  - name: Amy Browne
    affil: 1
  - name: Tsz Mang Yip
    affil: 2
affiliation:
  - num: 1
    address: University of Galway, Ireland
  - num: 2
    address: 
column_numbers: 3
output: 
  posterdown::posterdown_html:
    self_contained: false
bibliography: packages.bib
---

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(haven)
library(scales)
```

# Introduction

Missing data is often overlooked when analysing and reporting clinical trial results. There is little investigation conducted on the implications of simply ignoring missing data. We do know however that ignoring missing data can produce biased results and reduce statistical power. This is why its important to acknowledge the unobserved data and prioritize handling the data to increase precision and reduce bias. 

# Missingness Mechanisms

Missing Completely at Random (MCAR) - Missing data is independent of the study 
Missing at Random - Uncertainty as to how the data is missing
Missing Not at Random - Data informatively missing  

# Techniques to Handle Missing Data

1. Simple Imputation
2. Last Observation Carried Forward
3. Mean Imputation
4. Multiple Imputation
5. Missing Data Indicator 

# Visualising Missing Data

The following graphs visualise the frequency of missingness within two clinical trial data sets. 
The first set 

```{r mytable, echo=FALSE, fig.cap='Plot that shows the percentage of data thats unobserved', fig.height=9, fig.width=14}

# Formating data

data_vitalkp <- read_sas("Data/vitalkneepain111623.sas7bdat", 
                         NULL)
data_main <- read_sas("Data/VITAL_trial_NEJM_2022.sas7bdat", 
                      NULL)

data_combined <- data_main %>%
  select(Subject_ID, vitdactive, fishoilactive) %>%
  right_join(data_vitalkp, by = "Subject_ID")

data_miss <- data_combined %>%
  select(Subject_ID, tkrf, vitdactive, fishoilactive, starts_with("pain_")) %>%
  mutate(across(starts_with("pain_"), ~ as.numeric(!is.na(.x)), .names = "missing_{.col}")) %>%
  unite("miss_pattern", starts_with("missing_"), remove = FALSE) %>%
  rowwise() %>% 
  mutate(miss_count = sum(c_across(starts_with("missing_pain_")))) %>%
  group_by(miss_pattern) %>%
  mutate(pattern_freq = n()) %>%
  ungroup() %>%
  rowwise() %>%
  mutate(miss_pattern = gsub("_", ", ", miss_pattern),
         miss_pattern_label = paste0("Pattern: ", miss_pattern, ", (n=", pattern_freq , ")", collapse = "")) %>%
  pivot_longer(starts_with("pain_"), names_to = "time", names_prefix = "pain_", 
               values_to = "pain") %>%
  mutate(
    miss_obs = as.factor(if_else(is.na(pain),0,1)),
    time_cont = case_when(
      time=="base" ~ 0,
      time=="yr1" ~ 1,
      time=="yr2" ~ 2,
      time=="yr3" ~ 3,
      time=="yr4" ~ 4
    )) %>%
  arrange(desc(miss_pattern), Subject_ID) %>%
  mutate(miss_pattern_label = factor(miss_pattern_label, levels = unique(miss_pattern_label)),
         group_vitd = factor(vitdactive, levels = c(0,1), labels = c("Placebo","Vitamin D")),
         group_oil = factor(fishoilactive, levels = c(0,1), labels = c("Placebo","Fish Oil")))

## Ploting

pct_format = scales::label_percent(accuracy = .1)
suppressMessages(
data_miss %>%
  group_by(time, miss_obs) %>%
  summarise(count_pat = n(), .groups="keep") %>%
  group_by(time) %>%
  mutate(total_pat = sum(count_pat), perc_pat = pct_format(count_pat/total_pat)) %>%
  ungroup %>%
  ggplot(aes(x = time, fill = miss_obs, y = count_pat)) +
  geom_col() + 
  stat_identity(geom = "text", colour = "white", size = 3.5,
                aes(label =perc_pat), 
                position = position_stack(vjust=0.5)) +
  scale_fill_discrete(name = "Data Observed\nIndicator (R)", ) +
  scale_x_discrete(name = "Time") +
  scale_y_continuous(name = "Count (patients)") +
  theme(text = element_text(size = 18),
        axis.text.x = element_text(angle = 90, vjust = 0.5))
)
```

```{r mytable2, echo=FALSE, warning=FALSE, fig.cap='Plot that shows the percentage of data thats unobserved', fig.height=9, fig.width=15}

## Second plot
suppressMessages(
data_miss %>%
  group_by(time, miss_obs, group_vitd, group_oil) %>%
  summarise(count_pat = n(), .groups="keep") %>%
  group_by(time, group_vitd, group_oil) %>%
  mutate(total_pat = sum(count_pat), perc_pat = pct_format(count_pat/total_pat)) %>%
  ungroup %>%
  ggplot(aes(x = time, fill = miss_obs, y = count_pat)) +
  geom_col() + 
  facet_grid(group_vitd~group_oil) + 
  stat_identity(geom = "text", colour = "white", size = 3.5,
                aes(label =perc_pat), 
                position = position_stack(vjust=0.5)) +
  scale_fill_discrete(name = "Data Observed\nIndicator (R)", ) +
  scale_x_discrete(name = "Time") +
  scale_y_continuous(name = "Count (patients)") +
  theme(text = element_text(size = 18),
        axis.text.x = element_text(angle = 90, vjust = 0.5))
)
```

```{r mytable3, echo=FALSE, warning=FALSE, fig.cap='Plot that shows the percentage of data thats unobserved', fig.height=9, fig.width=14}

## Third plot
  ggplot(filter(data_miss, pattern_freq >= 10, miss_count > 2, time_cont > 0), aes(x = time_cont, y = pain)) +
  geom_point(aes(colour=group_oil), alpha = 0.2, size = 3, position = position_jitter(w = 0.1, h = 0)) +
  geom_smooth(aes(colour=group_oil, quiet=TRUE), method = "loess", linewidth = 2) +
  facet_wrap(~miss_pattern_label) +
  scale_x_continuous(name = "Time (year)" ) +
  scale_y_continuous(name = "Pain Score" ) +
  scale_colour_discrete(name = "Randomised Group" ) +
  theme(text = element_text(size = 18))

```

```{r mytable4, echo=FALSE, warning=FALSE, fig.cap='Plot that shows the percentage of data thats unobserved', fig.height=9, fig.width=14}
# Formating data

library(readr)
library(ggplot2)
acupuncture <- read_csv("Data/acupuncture.csv",show_col_types = FALSE)
acupuncture_miss <- acupuncture %>%
  select(id, group, pk1, pk2, pk5) %>%
  mutate(across(starts_with("pk"), ~ as.numeric(!is.na(.x)), .names = "missing_{.col}")) %>%
  unite("miss_pattern", starts_with("missing_"), remove = FALSE) %>%
  rowwise() %>% 
  mutate(miss_count = sum(c_across(starts_with("missing_pain_")))) %>%
  group_by(miss_pattern) %>%
  mutate(pattern_freq = n()) %>%
  ungroup() %>%
  rowwise() %>%
  mutate(miss_pattern = gsub("_", ", ", miss_pattern),
         miss_pattern_label = paste0("Pattern: ", miss_pattern, ", (n=", pattern_freq , ")", collapse = "")) %>%
  pivot_longer(starts_with("pk"), names_to = "time", names_prefix = "pk", 
               values_to = "pk") %>%
  mutate(
    miss_obs = as.factor(if_else(is.na(pk),0,1)),
    time_cont = case_when(
      time==1 ~ 0,
      time==2 ~ 1,
      time==5 ~ 2
    )) %>%
  arrange(desc(miss_pattern), id) %>%
  mutate(miss_pattern_label = factor(miss_pattern_label, levels = unique(miss_pattern_label)))

# 1st plot

pct_format = scales::label_percent(accuracy = .1)

acupuncture_miss %>%
  group_by(time, miss_obs) %>%
  summarise(count_pat = n()) %>%
  group_by(time) %>%
  mutate(total_pat = sum(count_pat), perc_pat = pct_format(count_pat/total_pat)) %>%
  ungroup %>%
  ggplot(aes(x = time, fill = miss_obs, y = count_pat)) +
  geom_col() + 
  stat_identity(geom = "text", colour = "white", size = 3.5,
                aes(label =perc_pat), 
                position = position_stack(vjust=0.5)) +
  scale_fill_discrete(name = "Data Observed\nIndicator (R)", ) +
  scale_x_discrete(name = "Time") +
  scale_y_continuous(name = "Count (patients)") +
  theme(text = element_text(size = 18),
        axis.text.x = element_text(angle = 90, vjust = 0.5))
```

```{r mytable5, echo=FALSE, warning=FALSE, fig.cap='Plot that shows the percentage of data thats unobserved', fig.height=9, fig.width=14}
# 2nd plot

acupuncture_miss %>%
  group_by(time, miss_obs, group) %>%
  summarise(count_pat = n()) %>%
  group_by(time, group) %>%
  mutate(total_pat = sum(count_pat), perc_pat = pct_format(count_pat/total_pat)) %>%
  ungroup %>%
  ggplot(aes(x = time, fill = miss_obs, y = count_pat)) +
  geom_col() + 
  facet_grid(.~group) + 
  stat_identity(geom = "text", colour = "white", size = 3.5,
                aes(label =perc_pat), 
                position = position_stack(vjust=0.5)) +
  scale_fill_discrete(name = "Data Observed\nIndicator (R)", ) +
  scale_x_discrete(name = "Time") +
  scale_y_continuous(name = "Count (patients)") +
  theme(text = element_text(size = 18),
        axis.text.x = element_text(angle = 90, vjust = 0.5))
```





# Prediction

# Conclusion

Try `posterdown` out! Hopefully you like it!

```{r, include=FALSE}
knitr::write_bib(c('knitr','rmarkdown','posterdown','pagedown'), 'packages.bib')
```

# References

