---
title: Estimating Treatment Effects in Longitudinal Clinical Trials with Missing Data
author:
  - name: Amy Browne
  - name: Tsz Mang Yip
output:
   ioslides_presentation:
    css: styles.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(mice)
library(tidyverse)
library(haven)
library(lme4)
library(broom.mixed)
library(meta)
library(dplyr)
library(naniar)
library(tidymodels)
library(gtsummary)
source("R scripts/Functions.R")
source("R scripts/VITAL data formats.R")
source("R scripts/Acupuncture data formats.R")
source("R scripts/Summary analysis.R")
source("R scripts/Category time analysis.R")
source("R scripts/Continuous time analysis.R")
source("R scripts/Different Imputation Method.R")
source("R scripts/Sensitive Analysis.R")
```

# Recapitulation

# Project Background 

## Missing Data

Missing data is defined as data that has not been stored for a variable in the observation of interest. A systematic review of articles regarding 62 studies found that:
- 44% did not report missing data 
- 69% did not report how missing data was handled

This shows how missing data is a topic that is often negected even though it can cause biased results when hadled inappropriately. 

## Missingness Mechanisms 

- Missing Completely At Random: $$P(R = 1 \mid Y, X) = P(R = 1)$$
- Missing At Random: $$P(R = 1 \mid Y, X) = P(R = 1 \mid X)$$
- Missing Not At Random: $$P(R = 1 \mid Y, X) \ne P(R = 1 \mid X)$$
- We can never confirm these assumptions but Little's MCAR Test can rule out the MCAR assumption

# Our Datasets 

## Treatment of Acupuncture on Headache 

- Randomised controlled trial to determine the treatment effect of acupuncture on chronic headache against a placebo

```{r}
acu_summary
```


## Treatment of Vitamin D and Fish Oil on Knee Pain

# Handling Missing Data 

## Complete Case Analysis 

- Only analyse the cases where there are no missing data in any variables  
- May be valid when data is MCAR only
- Introduces selection bias 
- Reduces statistical power 

## Single Imputation

- Replaces missing data with a single value 
- Last observation carried forward
- Mean observation carried forward 
- Reduces variability in the results 

## Maximum Likelihood

- Linear Mixed Effects Model 

## Multiple Imputation Using Chained Equations 

- All missing data are imputed with values derived from observed data to act as placeholders
- Starting with variable $x_i$ that contains missing data, remove the place holders in this variable 
- Regress $x_i$ on remaining covariates $x_{i+1}$, $x_{i+2}$,..., $x_{i+p}$
- This model is predicts the missing values in $x_i$ if it were observed, these steps are repeated for $x_{i+1}$, $x_{i+2}$,..., $x_{i+p}$
- We can use this process to create $m$ data sets with imputed values, typically 5 or 10 suffices. 

## MICE function

| Method         | Definition                              |
|----------------|-----------------------------------------|
| "pmm"          | Predictive mean matching                |
| "norm.nob"     | Linear regression, ignoring model error |
| "norm"         | Bayesian linear regression              |
| "norm.predict" | Linear regression, predicted value      |
| "sample"       | Random sample from $Y_O$                |
| "midastouch"   | Weighted predictive mean matching       |



## Estimating Treatment Effects 

- Simple Linear Regression Models: Outcome is modelled as a straight line 
- Linear Mixed Effects Models: Extends the linear regression model by accounting for individual variability

## Treatment Effect of Acupuncture 

## Continous 

Linear Regression Model 1
$$\text{pk_score} = \beta_0 + \beta_1 \cdot \text{group} + \beta_2 \cdot \text{time} + \beta_3 \cdot (\text{group} \cdot \text{time}) + \beta_4 \cdot \text{pk1} + \varepsilon$$

Linear Mixed Effects model 1
$$\text{pk_score}_{ij} = \beta_0 + \beta_1 \cdot \text{group}_i + \beta_2 \cdot \text{time}_{ij} + \beta_3 \cdot (\text{group}_i \cdot \text{time}_{ij}) + \beta_4 \cdot \text{pk1}_i + b_{0i} + \varepsilon_{ij}$$

## Categorical

Linear Regression Model 2
$$\text{pk5} = \beta_0 + \beta_1 \cdot \text{group} + \beta_2 \cdot \text{pk1} + \varepsilon$$

Linear Mixed Effects 2
$$\text{pk5}_{ij} = \beta_0 + \beta_1 \cdot \text{group}_i + \beta_2 \cdot \text{pk1}_i + \beta_3 \cdot \text{pk2}_{ij} + b_{0i} + \varepsilon_{ij}$$

## Acupuncture Results 


```{r echo=TRUE}
acu_plot_categorical
```

## Comparing when time is continous 

```{r echo=TRUE}
acu_plot_compare
```

