---
title: Estimating Treatment Effects in Clinical Trials with Missing Data
author:
  - name: Amy Browne
  - name: Tsz Mang Yip
    affil: 1
affiliation:
  - num: 1
    address: School of Mathematical and Statistical Sciences, University of Galway

poster_height: "594mm"
poster_width: "841mm"
column_numbers: 4
column_margins: "0.5in"	
sectitle_textsize: "40pt"
primary_colour:	"#33CCFF"	
secondary_colour:	"#6699FF"	
accent_colour:	"#000033"	
author_textcol: "white"
title_textsize: "60pt"
author_textsize: "40pt"
body_textsize: "20pt"
output: 
  posterdown::posterdown_html:
    self_contained: true
    pandoc_args: --mathjax
    number_sections: false
knit: pagedown::chrome_print
---

```{css, echo=FALSE}
div.logo_left{
  width: 20%;
}
div.poster_title{
  width: 80%;
}
.section h4 {
    break-after: column;
}
```

<!-- Don't change anything above, except the title and author names, unless you know what you are doing. -->

```{r, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      tidy = FALSE,
                      message = FALSE,
                      fig.align = 'center',
                      out.width = "100%")
options(knitr.table.format = "html") 
# Load any additional libraries here
library(tidyverse)
library(plotly)
library(kableExtra)
library(haven)
library(scales)
```

# Outline

Missing data is often overlooked when analysing and reporting clinical trial results. There is little investigation conducted on the implications of simply ignoring missing data. We do know however that ignoring missing data can produce biased results and reduce statistical power. This is why its important to acknowledge the unobserved data and prioritize handling the data to increase precision and reduce bias. 

# Missingness Mechanisms

- Missing Completely at Random (MCAR) - Missing data is independent of the study 
- Missing at Random - Uncertainty as to how the data is missing
- Missing Not at Random - Data is informatively missing  

# Our Data

The following graphs visualise the frequency of missingness within two clinical trial data sets. The first data set contains results of a longitudinal Randomised Control Trial which aimed to test the effectiveness of in taking Fish oil or Vitamin D on Knee Pain. 
The following data set are results from another longitudinal randomised control trial in which the effectiveness of acupuncture on chronic headache was investigated.

# Vitamin D and Fish Oil Trial

```{r}

data_vitalkp <- read_sas("Data/vitalkneepain111623.sas7bdat", 
                         NULL)
data_main <- read_sas("Data/VITAL_trial_NEJM_2022.sas7bdat", 
                      NULL)

data_combined <- data_main %>%
  select(Subject_ID, vitdactive, fishoilactive) %>%
  right_join(data_vitalkp, by = "Subject_ID")

data_miss <- data_combined %>%
  select(Subject_ID, tkrf, vitdactive, fishoilactive, starts_with("pain_")) %>%
  mutate(across(starts_with("pain_"), ~ as.numeric(!is.na(.x)), .names = "missing_{.col}")) %>%
  unite("miss_pattern", starts_with("missing_"), remove = FALSE) %>%
  rowwise() %>% 
  mutate(miss_count = sum(c_across(starts_with("missing_pain_")))) %>%
  group_by(miss_pattern) %>%
  mutate(pattern_freq = n()) %>%
  ungroup() %>%
  rowwise() %>%
  mutate(miss_pattern = gsub("_", ", ", miss_pattern),
         miss_pattern_label = paste0("Pattern: ", miss_pattern, ", (n=", pattern_freq , ")", collapse = "")) %>%
  pivot_longer(starts_with("pain_"), names_to = "time", names_prefix = "pain_", 
               values_to = "pain") %>%
  mutate(
    miss_obs = as.factor(if_else(is.na(pain),0,1)),
    time_cont = case_when(
      time=="base" ~ 0,
      time=="yr1" ~ 1,
      time=="yr2" ~ 2,
      time=="yr3" ~ 3,
      time=="yr4" ~ 4
    )) %>%
  arrange(desc(miss_pattern), Subject_ID) %>%
  mutate(miss_pattern_label = factor(miss_pattern_label, levels = unique(miss_pattern_label)),
         group_vitd = factor(vitdactive, levels = c(0,1), labels = c("Placebo","Vitamin D")),
         group_oil = factor(fishoilactive, levels = c(0,1), labels = c("Placebo","Fish Oil")))
```

```{r mytable, echo=FALSE, fig.cap='', fig.height=9, fig.width=14}
## Ploting

pct_format = scales::label_percent(accuracy = .1)
suppressMessages(
data_miss %>%
  group_by(time, miss_obs) %>%
  summarise(count_pat = n(), .groups="keep") %>%
  group_by(time) %>%
  mutate(total_pat = sum(count_pat), perc_pat = pct_format(count_pat/total_pat)) %>%
  ungroup %>%
  ggplot(aes(x = time, fill = miss_obs, y = count_pat)) +
  geom_col() + ggtitle("Pattern of Missing Data over time in Vitamin D & Fish Oil trial") + 
  stat_identity(geom = "text", colour = "white", size = 10,
                aes(label =perc_pat), 
                position = position_stack(vjust=0.5)) +
  scale_fill_manual(name = "Data Observed", values=c("mediumpurple1", "slateblue3")) +
  scale_x_discrete(name = "Time (years)") +
  scale_y_continuous(name = "Number of Patients") +
  theme(text = element_text(size = 18),
        axis.text.x = element_text(angle = 90, vjust = 0.5),
        legend.position = "top")
)
```
```{r mytable2, echo=FALSE, warning=FALSE, fig.cap='Data Observed (dark purple), Missing Data (light purple). We can see that proportions are similar within groups', fig.height=9, fig.width=15}

## Second plot
suppressMessages(
data_miss %>%
  group_by(time, miss_obs, group_vitd, group_oil) %>%
  summarise(count_pat = n(), .groups="keep") %>%
  group_by(time, group_vitd, group_oil) %>%
  mutate(total_pat = sum(count_pat), perc_pat = pct_format(count_pat/total_pat)) %>%
  ungroup %>%
  ggplot(aes(x = time, fill = miss_obs, y = count_pat)) +
  geom_col() + ggtitle("Proportions of missing data from baseline to year 4 within treatment groups") +
  facet_grid(group_vitd~group_oil) + 
  stat_identity(geom = "text", colour = "white", size = 10,
                aes(label =perc_pat), 
                position = position_stack(vjust=0.5)) +
  scale_fill_manual(values=c("mediumpurple1", "slateblue3")) +
  scale_x_discrete(name = "Time") +
  scale_y_continuous(name = "Number of Patients") +
  theme(text = element_text(size = 18),
        axis.text.x = element_text(angle = 90, vjust = 0.5), legend.position = "none")
)
```
 
```{r mytable3, echo=FALSE, warning=FALSE, message=FALSE, fig.cap='', fig.height=9, fig.width=14}

## Third plot
  ggplot(filter(data_miss, pattern_freq >= 10, miss_count > 2, time_cont > 0), aes(x = time_cont, y = pain)) + ggtitle("Missing data pattern over time") +
  geom_point(aes(colour=group_oil), alpha = 0.2, size = 3, position = position_jitter(w = 0.1, h = 0)) +
  geom_smooth(aes(colour=group_oil), method = "loess", linewidth = 2) +
  facet_wrap(~miss_pattern_label) +
  scale_x_continuous(name = "Time (year)" ) +
  scale_y_continuous(name = "Pain Score" )  +
  scale_colour_manual(name="Treatment Group",values=c("blue", "yellow")) +
  theme(text = element_text(size = 18), legend.position="top")

```

#### 

# Acupuncture Trial

```{r}
# Formating data

library(readr)
library(ggplot2)
acupuncture <- read_csv("Data/acupuncture.csv",show_col_types = FALSE)
acupuncture_miss <- acupuncture %>%
  select(id, group, pk1, pk2, pk5) %>%
  mutate(across(starts_with("pk"), ~ as.numeric(!is.na(.x)), .names = "missing_{.col}")) %>%
  unite("miss_pattern", starts_with("missing_"), remove = FALSE) %>%
  rowwise() %>% 
  mutate(miss_count = sum(c_across(starts_with("missing_pain_")))) %>%
  group_by(miss_pattern) %>%
  mutate(pattern_freq = n()) %>%
  ungroup() %>%
  rowwise() %>%
  mutate(miss_pattern = gsub("_", ", ", miss_pattern),
         miss_pattern_label = paste0("Pattern: ", miss_pattern, ", (n=", pattern_freq , ")", collapse = "")) %>%
  pivot_longer(starts_with("pk"), names_to = "time", names_prefix = "pk", 
               values_to = "pk") %>%
  mutate(
    miss_obs = as.factor(if_else(is.na(pk),0,1)),
    time_cont = case_when(
      time==1 ~ 0,
      time==2 ~ 1,
      time==5 ~ 2
    )) %>%
  arrange(desc(miss_pattern), id) %>%
  mutate(miss_pattern_label = factor(miss_pattern_label, levels = unique(miss_pattern_label)))
```
```{r mytable4, echo=FALSE, warning=FALSE, fig.cap='1-Baseline, 2- 3 Months, 5- 1 Year', fig.height=9, fig.width=14}
# 1st plot

pct_format = scales::label_percent(accuracy = .1)

acupuncture_miss %>%
  group_by(time, miss_obs) %>%
  summarise(count_pat = n(), .groups = "keep") %>%
  group_by(time) %>%
  mutate(total_pat = sum(count_pat), perc_pat = pct_format(count_pat/total_pat)) %>%
  ungroup %>%
  ggplot(aes(x = time, fill = miss_obs, y = count_pat)) +
  geom_col() + ggtitle("Proportions of Missing Data over time") +
  stat_identity(geom = "text", colour = "white", size = 10,
                aes(label =perc_pat), 
                position = position_stack(vjust=0.5)) +
  scale_fill_manual(name="Data Observed",values=c("lightsteelblue","royalblue")) + 
  scale_x_discrete(name = "Time") +
  scale_y_continuous(name = "Count (patients)") +
  theme(text = element_text(size = 18),
        axis.text.x = element_text(angle = 90, vjust = 0.5), legend.position = "top")
```

```{r mytable5, echo=FALSE, warning=FALSE, fig.cap='Light Blue = Missing, Darker Blue = Observed', fig.height=9, fig.width=14}
# 2nd plot

acupuncture_miss %>%
  group_by(time, miss_obs, group) %>%
  summarise(count_pat = n(), .groups="keep") %>%
  group_by(time, group) %>%
  mutate(total_pat = sum(count_pat), perc_pat = pct_format(count_pat/total_pat)) %>%
  ungroup %>%
  ggplot(aes(x = time, fill = miss_obs, y = count_pat)) + ggtitle("Proportions of Missing Data within Treatment Groups over time (0=Control, 1=Acupuncture") +
  geom_col() + 
  facet_grid(.~group) + 
  stat_identity(geom = "text", colour = "white", size = 10,
                aes(label =perc_pat), 
                position = position_stack(vjust=0.5)) +
  scale_fill_manual(name = "Data Observed", values=c("lightsteelblue","royalblue")) + 
  scale_x_discrete(name = "Time") +
  scale_y_continuous(name = "Count (patients)") +
  theme(text = element_text(size = 18),
        axis.text.x = element_text(angle = 90, vjust = 0.5),legend.position = "none")
```


####

# Data Handling Techniques
1. Last Observation Carried Forward; Replacing the missing data with the last value that was observed [^1]
2. Simple Imputation; Replacing the missing data with a calcualted statistic of the observed data (such as mean or median)
3. Indicator Method; Using a statistical model of the observations, a dummy variable is created to indicate missing or observed.
4. Multiple Imputation; The observed data is used to create a distribution, from which K values are drawn to form K data sets. Each data set is analysed then combined to form an overall effect. 

# GitHub

The code and datasets for this project can be viewed at our GitHub repository here: <https://github.com/>

# References

[^1]: Carpenter & Kenward, Missing data in randomised controlled trials — a practical guide
