---
title: Estimating Treatment Effects in Longitudinal Clinical Trials with Missing Data
author:
  - name: Amy Browne
  - name: Tsz Mang Yip
    affil: 
affiliation:
  - num: 
    address: School of Mathematical and Statistical Sciences, University of Galway
logoleft_name: https&#58;//sport.universityofgalway.ie/assets/img/logo2.png
poster_height: "594mm"
poster_width: "841mm"
column_numbers: 4
font_family: "Arial"
column_margins: "0.3in"	
sectitle_textsize: "35pt"
primary_colour:	"#a80050"
secondary_colour:	"#84003d"
accent_colour:	"#000033"	
author_textcol: "white"
title_textsize: "55pt"
author_textsize: "35pt"
body_textsize: "20pt"
output: 
  posterdown::posterdown_html:
    self_contained: true
    pandoc_args: --mathjax
    number_sections: false
knit: pagedown::chrome_print
---

```{css, echo=FALSE}
div.logo_left{
  width: 20%;
}
div.poster_title{
  width: 80%;
}
.section h4 {
    break-after: column;
}
div.footnotes {
    font-size: 10pt;
}
table {
 font-size: 13pt;
}

```

<!-- Don't change anything above, except the title and author names, unless you know what you are doing. -->

```{r, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      tidy = FALSE,
                      message = FALSE,
                      fig.align = 'center',
                      out.width = "100%")
options(knitr.table.format = "html") 
# Load any additional libraries here
library(tidyverse)
library(plotly)
library(kableExtra)
library(haven)
library(scales)
```

# Outline

Missing data is often overlooked when analysing and reporting clinical trial results. Excluding observations on the basis of missing data can produce biased results and reduce statistical power. It is important to plan for and describe the occurrence of missing data and consider how such missing data should be incorporated in any analysis estimating a treatment in order to increase precision and reduce bias. A key statistical consideration in longitudinal trials is the non-independence of repeated observations from the same patient, which must be accounted for in any analysis.

# Missingness Mechanisms

To account for missing data, it must be understood the causes of data being missing. There are three mechanisms of missing data that when applied will decide which data handling methods are valid to use. 

- Missing Completely at Random (MCAR) - Missing data is unrelated to making inference about the treatment effect. This is the best case scenario as the results remain unbiased and the intervention effect unchanged.  
- Missing at Random - Missing data is systematically related to the observed variables but not the unobserved variables. [^5]
- Missing Not at Random - Missing data is related to the unobserved data itself. We must investigate the data distributions of patients with missing data and compare to full data distributions. 

# Project Aims

This project investigates missing data analysis methods in two longitudinal RCT data sets. The trials are summarised in the following table:

| Trial                  | Vitamin D & omega trial[^1]  | Acupuncture Trial[^2]       |
| -----------------------| -----------------------------| ----------------------------|
| Sample Size            | 1398                         | 401                         |
| Trial                  | 2x2 Longitudinal RCT         | Longitudinal RCT            |
| Follow-up              | 1 year intervals for 5 years | Baseline, 3 months, 1 year  |
| Independent Variable(s)| Vitamin D, Fish Oil          | Acupuncture treatment       |
| Dependent Variable(s)  | Knee pain score              | Migraine severity score     |

####

# Vitamin D and Fish Oil Trial [^3]

```{r}

data_vitalkp <- read_sas("Data/vitalkneepain111623.sas7bdat", 
                         NULL)
data_main <- read_sas("Data/VITAL_trial_NEJM_2022.sas7bdat", 
                      NULL)

data_combined <- data_main %>%
  select(Subject_ID, vitdactive, fishoilactive) %>%
  right_join(data_vitalkp, by = "Subject_ID")

data_miss <- data_combined %>%
  select(Subject_ID, tkrf, vitdactive, fishoilactive, starts_with("pain_")) %>%
  mutate(across(starts_with("pain_"), ~ as.numeric(!is.na(.x)), .names = "missing_{.col}")) %>%
  unite("miss_pattern", starts_with("missing_"), remove = FALSE) %>%
  rowwise() %>% 
  mutate(miss_count = sum(c_across(starts_with("missing_pain_")))) %>%
  group_by(miss_pattern) %>%
  mutate(pattern_freq = n()) %>%
  ungroup() %>%
  rowwise() %>%
  mutate(miss_pattern = gsub("_", ", ", miss_pattern),
         miss_pattern_label = paste0("Pattern: ", miss_pattern, ", (n=", pattern_freq , ")", collapse = "")) %>%
  pivot_longer(starts_with("pain_"), names_to = "time", names_prefix = "pain_", 
               values_to = "pain") %>%
  mutate(
    miss_obs = as.factor(if_else(is.na(pain),0,1)),
    time_cont = case_when(
      time=="base" ~ 0,
      time=="yr1" ~ 1,
      time=="yr2" ~ 2,
      time=="yr3" ~ 3,
      time=="yr4" ~ 4
    )) %>%
  arrange(desc(miss_pattern), Subject_ID) %>%
  mutate(miss_pattern_label = factor(miss_pattern_label, levels = unique(miss_pattern_label)),
         group_vitd = factor(vitdactive, levels = c(0,1), labels = c("Placebo","Vitamin D")),
         group_oil = factor(fishoilactive, levels = c(0,1), labels = c("Placebo","Fish Oil")))
```

```{r mytable, echo=FALSE, fig.height=5, fig.width=14}
pct_format = scales::label_percent(accuracy = .1)
suppressMessages(
data_miss %>%
  group_by(time, miss_obs) %>%
  summarise(count_pat = n(), .groups="keep") %>%
  group_by(time) %>%
  mutate(total_pat = sum(count_pat), perc_pat = pct_format(count_pat/total_pat)) %>%
  ungroup %>%
  ggplot(aes(x = time, fill = miss_obs, y = count_pat)) +
  geom_col() +  
  stat_identity(geom = "text", colour = "white", size = 8,
                aes(label =perc_pat), 
                position = position_stack(vjust=0.5)) +
scale_fill_manual(name="",values=c("olivedrab", "lawngreen"), labels=c("Missing Data", "Measured Data")) +
  scale_x_discrete(name = "Time", labels = c("base" = "Baseline", "yr1" = "Year 1", "yr2" = "Year 2", "yr3" = "Year 3", "yr4"="Year 4")) + 
  scale_y_continuous(name = "Number of Patients") +
  theme(text = element_text(size = 24),
        axis.text.x = element_text(angle = 0, vjust = 0.5),
        legend.position = "top", legend.text=element_text(size=20))
)
```
Figure 1. The proportions of missing data (represented in dark green) at different time points. The observed missing data pattern is non-monotonic.

```{r mytable2, echo=FALSE, warning=FALSE, fig.height=6, fig.width=14}
suppressMessages(
data_miss %>%
  group_by(time, miss_obs, group_oil) %>%
  summarise(count_pat = n(), .groups="keep") %>%
  group_by(time, group_oil) %>%
  mutate(total_pat = sum(count_pat), perc_pat = pct_format(count_pat/total_pat)) %>%
  ungroup %>%
  ggplot(aes(x = time, fill = miss_obs, y = count_pat)) +
  geom_col() +
  facet_grid(~group_oil) + 
  stat_identity(geom = "text", colour = "white", size = 5,
                aes(label =perc_pat), 
                position = position_stack(vjust=0.5)) +
  scale_fill_manual(name="",values=c("olivedrab", "lawngreen"))+
  scale_x_discrete(name = "Time", labels = c("base" = "Baseline", "yr1" = "Year 1", "yr2" = "Year 2", "yr3" = "Year 3", "yr4"="Year 4")) + 
  scale_y_continuous(name = "Number of Patients") +
  theme(text = element_text(size = 24),
        axis.text.x = element_text(angle = 0, vjust = 0.5), legend.position = "none")
)
```
 
Figure 2. Proportions of missing data (dark green) over time in different groups (Fish Oil vs. Placebo) are shown. Similar proportions of missing data were observed between the two groups.
 
```{r mytable3, echo=FALSE, warning=FALSE, message=FALSE, fig.height=9, fig.width=14}
data_miss_count <- data_miss %>%
  select(Subject_ID, miss_pattern, pain, time_cont, group_oil, miss_obs) %>%
  group_by(group_oil, miss_pattern, miss_obs, time_cont) %>%
  summarise(pattern_count = n()) %>%
  ungroup %>%
  mutate(pattern_perc = round(pattern_count/sum(pattern_count)*100*5,1))

ggplot(data_miss_count, aes(x = time_cont, y = miss_pattern, fill = miss_obs)) +
  geom_tile(color = "white") + 
  geom_text(data = filter(data_miss_count, time_cont == 4), 
            aes(x = time_cont+1.5, label = paste0(pattern_count," (", pattern_perc, "%)")), size = 4) +
  facet_grid(~group_oil, scales = "free", switch = "y") +
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.spacing.x = unit(1, "lines"),
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 90, vjust = 0.5) , legend.position = "none") +
  coord_cartesian(xlim = c(-0.5, 6.2), clip = 'off') +
  scale_y_discrete(name = "Missing Data Pattern") +
  scale_fill_discrete(name = "Data Observed\nIndicator (R)", ) +
  scale_x_continuous(name = "Time", breaks = c(0,1,2,3,4,5), 
                     labels = c("Baseline","Year 1","Year 2","Year 3","Year 4","Patient Count\n(%)")) +
  scale_fill_manual(name="",labels=c("Missing Data","Measured Data"),values=c("olivedrab", "lawngreen"))
```

Figure 3. The missing data patterns (where 0 represents missing data and 1 represents observed data) between the groups (Fish Oil vs. Placebo) are displayed. Some differences in the patterns can be observed.

####

# Acupuncture Trial

```{r}
library(readr)
library(ggplot2)
acupuncture <- read_csv("Data/acupuncture.csv",show_col_types = FALSE)
acupuncture_miss <- acupuncture %>%
  select(id, group, pk1, pk2, pk5) %>%
  mutate(across(starts_with("pk"), ~ as.numeric(!is.na(.x)), .names = "missing_{.col}")) %>%
  unite("miss_pattern", starts_with("missing_"), remove = FALSE) %>%
  rowwise() %>% 
  mutate(miss_count = sum(c_across(starts_with("missing_pain_")))) %>%
  group_by(miss_pattern) %>%
  mutate(pattern_freq = n()) %>%
  ungroup() %>%
  rowwise() %>%
  mutate(miss_pattern = gsub("_", ", ", miss_pattern),
         miss_pattern_label = paste0("Pattern: ", miss_pattern, ", (n=", pattern_freq , ")", collapse = "")) %>%
  pivot_longer(starts_with("pk"), names_to = "time", names_prefix = "pk", 
               values_to = "pk") %>%
  mutate(
    miss_obs = as.factor(if_else(is.na(pk),0,1)),
    time_cont = case_when(
      time==1 ~ 0,
      time==2 ~ 1,
      time==5 ~ 2
    )) %>%
  arrange(desc(miss_pattern), id) %>%
  mutate(miss_pattern_label = factor(miss_pattern_label, levels = unique(miss_pattern_label)))
```

```{r mytable4, echo=FALSE, warning=FALSE, fig.height=7, fig.width=14}
pct_format = scales::label_percent(accuracy = .1)

acupuncture_miss %>%
  group_by(time, miss_obs) %>%
  summarise(count_pat = n(), .groups = "keep") %>%
  group_by(time) %>%
  mutate(total_pat = sum(count_pat), perc_pat = pct_format(count_pat/total_pat)) %>%
  ungroup %>%
  ggplot(aes(x = time, fill = miss_obs, y = count_pat)) +
  geom_col() +
  stat_identity(geom = "text", colour = "white", size = 10,
                aes(label =perc_pat), 
                position = position_stack(vjust=0.5)) +
  scale_fill_manual(name="",values=c("olivedrab", "lawngreen"), labels=c("Missing Data", "Measured Data")) + 
  scale_x_discrete(name = "Time", labels = c("1" = "Baseline", "2" = "3 Months", "5" = "12 Months")) +
  scale_y_continuous(name = "Number of Patients") +
  theme(text = element_text(size = 18),
        axis.text.x = element_text(angle = 0, vjust = 0.5), legend.position = "top", legend.text=element_text(size=20))
```

Figure 4. Proportions of Missing data (dark green) at different time in Acupuncture trial 

```{r mytable5, echo=FALSE, warning=FALSE, fig.height=8, fig.width=14}
acupuncture_miss %>%
  group_by(time, miss_obs, group) %>%
  summarise(count_pat = n(), .groups="keep") %>%
  group_by(time, group) %>%
  mutate(total_pat = sum(count_pat), perc_pat = pct_format(count_pat/total_pat)) %>%
  ungroup %>%
  ggplot(aes(x = time, fill = miss_obs, y = count_pat)) +
  geom_col() + 
  facet_grid(.~group, labeller=labeller(group=c("0" = "Placebo", "1" = "Acupuncture"))) + 
  stat_identity(geom = "text", colour = "white", size = 10,
                aes(label =perc_pat), 
                position = position_stack(vjust=0.5)) +
  scale_fill_manual(name="Data",values=c("olivedrab", "lawngreen"), labels=c("Missing", "Measured")) + 
 scale_x_discrete(name = "Time", labels = c("1" = "Baseline", "2" = "3 Months", "5" = "12 Months")) +
  scale_y_continuous(name = "Number of Patients") +
  theme(text = element_text(size = 18),
        axis.text.x = element_text(angle = 0, vjust = 0.5),legend.position = "none",strip.text.x=element_text(size=20))
```

Figure 5. Proportions of missing data (dark green) within treatment groups over time. They are similar within both treatment groups, but higher percentage of data is missing in placebo group. 

```{r mytable6, echo=FALSE, warning=FALSE, fig.height=7, fig.width=14}
acupuncture_miss_count <- acupuncture_miss %>%
  select(id, miss_pattern, pk, time_cont, group, miss_obs) %>%
  group_by(group, miss_pattern, miss_obs, time_cont) %>%
  summarise(pattern_count = n(), .groups="keep") %>%
  ungroup %>%
  mutate(pattern_perc = round(pattern_count/sum(pattern_count)*100*5,1))


ggplot(acupuncture_miss_count, aes(x = time_cont, y = miss_pattern, fill = miss_obs)) +
  geom_tile(color = "white") + 
  geom_text(data = filter(acupuncture_miss_count, time_cont == 2), 
            aes(x = time_cont+1, label = paste0(pattern_count," (", pattern_perc, "%)")), size = 5.5) +
  facet_grid(~group, scales = "free", switch = "y",labeller=labeller(group=c("0" = "Placebo", "1" = "Acupuncture"))) +
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.spacing.x = unit(1, "lines"),
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 0, vjust = 0.5), legend.position = "none",strip.text.x=element_text(size=20)) +
  coord_cartesian(xlim = c(-0.5, 3.2), clip = 'off') +
  scale_y_discrete(name = "Missing Data Pattern") +
  scale_fill_manual(name = "", labels=c("Missing","Measured"), values=c("olivedrab","lawngreen")) +
  scale_x_continuous(name = "Time", breaks = c(0.1,1.1,2.1,3.1), 
                     labels = c("Baseline","3 Months","12 Months","Patients \n(%)")) 
```

Figure 6. Missing data pattern (where 0 represents missing data and 1 represents observed data) over time. Slight differences was observed.

####

# Common Analysis For Data With Missing Observations [^4]

1. Complete Case Analysis: Only includes participants with no missing data in the analysis, ignoring cases with any missing values.
2. Simple Imputation: Replaces missing values with a single estimate, such as Last/baseline observation carried forward (LOCF/BOCF).
3. Weighted Estimation: Applies weights to account for the likelihood of data being missing, such as inverse probability weighting (IPW).
4. Mix-effect Model: Uses likelihood-based methods to estimate parameters while accounting for missing data.
5. Multiple Imputation: Replaces missing data with multiple plausible values, generating multiple datasets for analysis and combining the result for inference.

# Next step

- We will use a mixed-effects model and multiple imputation (MI) as the primary methods for analyzing both datasets.
- Both methods assume the data are Missing at Random (MAR). A sensitivity analysis will be conducted to test this assumption.
- Mean imputation methods like LOCF/WOCF/BOCF are commonly used, often ad-hoc with unclear assumptions about the missing data mechanism, and tend to result in incorrect inference regarding quantities or parameters of interest. However, we will also conduct analysis in this way to see the impact on our dataset.

# References
[^1]: MacFarlane, L. A. et al., (2020). The Effects of Vitamin D and Marine Omega-3 Fatty Acid Supplementation on Chronic Knee Pain in Older US Adults: Results from a Randomized Trial. Arthritis & rheumatology (Hoboken, N.J.), 72(11), 1836–1844. https://doi.org/10.1002/art.41416
[^2]: Vickers, A. J. et al., (2004). Acupuncture for chronic headache in primary care: large, pragmatic, randomised trial. BMJ (Clinical research ed.), 328(7442), 744. https://doi.org/10.1136/bmj.38029.421863.EB
[^3]: GitHub: The code and datasets for this project can be viewed at our GitHub repository here: <https://github.com/amydebrun/Missing-Data.git>
[^4]: Carpenter, J., & Kenward, M. (2007). Missing data in randomised controlled trials: a practical guide. Health Technology Assessment Methodology Programme. 
[^5]: https://www.ncbi.nlm.nih.gov/books/NBK493614/