---
title: Estimating Treatment Effects in Longitudinal Clinical Trials with Missing Data
author:
  - name: Amy Browne
  - name: Tsz Mang Yip
    affil: 1
affiliation:
  - num: 1
    address: School of Mathematical and Statistical Sciences, University of Galway

poster_height: "594mm"
poster_width: "841mm"
column_numbers: 4
font_family: "Arial"
column_margins: "0.3in"	
sectitle_textsize: "35pt"
primary_colour:	"#a80050"
secondary_colour:	"#84003d"
accent_colour:	"#000033"	
author_textcol: "white"
title_textsize: "55pt"
author_textsize: "35pt"
body_textsize: "20pt"
output: 
  posterdown::posterdown_html:
    self_contained: true
    pandoc_args: --mathjax
    number_sections: false
knit: pagedown::chrome_print
---

```{css, echo=FALSE}
div.logo_left{
  width: 20%;
}
div.poster_title{
  width: 80%;
}
.section h4 {
    break-after: column;
}
div.footnotes {
    font-size: 10pt;
}
```

<!-- Don't change anything above, except the title and author names, unless you know what you are doing. -->

```{r, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      tidy = FALSE,
                      message = FALSE,
                      fig.align = 'center',
                      out.width = "100%")
options(knitr.table.format = "html") 
# Load any additional libraries here
library(tidyverse)
library(plotly)
library(kableExtra)
library(haven)
library(scales)
```

# Outline

Missing data is often overlooked when analysing and reporting clinical trial results. Excluding observations on the basis of missing data can produce biased results and reduce statistical power. It is important to plan for and describe the occurrence of missing data and consider how such missing data should be incorporated in any analysis estimating a treatment in order to increase precision and reduce bias. A key statistical consideration in longitudinal trials is the non-independence of repeated observations from the same patient, which must be accounted for in any analysis.

# Missingness Mechanisms

- Missing Completely at Random (MCAR) - Missing data is independent of the study 
- Missing at Random - Uncertainty as to how the data is missing
- Missing Not at Random - Data is informatively missing  

# Project Aims
This project investigates missing data analysis methods in two longitudinal RCT datasets. As the first part of the project, we are visualizing the proportion and patterns of missing values for each dataset.

The first dataset is from a 2x2 factorial RCT testing the effects of Fish Oil and Vitamin D on knee pain and their role in preventing cancer and cardiovascular disease (MacFarlane et al., 2020) [^1]. Over 20,000 healthy participants were randomized into four treatment groups, followed at one-year intervals for five years, with the primary endpoint being cardiovascular events.

The second dataset comes from an RCT evaluating acupuncture's effectiveness on chronic headaches, mainly migraines (Vickers et al., 2004) [^2]. A total of 401 patients were randomized to receive up to 12 acupuncture treatments over three months, with outcomes such as headache score, SF-36 health status, and medication use measured at baseline, three months, and 12 months.

####

# Vitamin D and Fish Oil Trial

```{r}

data_vitalkp <- read_sas("Data/vitalkneepain111623.sas7bdat", 
                         NULL)
data_main <- read_sas("Data/VITAL_trial_NEJM_2022.sas7bdat", 
                      NULL)

data_combined <- data_main %>%
  select(Subject_ID, vitdactive, fishoilactive) %>%
  right_join(data_vitalkp, by = "Subject_ID")

data_miss <- data_combined %>%
  select(Subject_ID, tkrf, vitdactive, fishoilactive, starts_with("pain_")) %>%
  mutate(across(starts_with("pain_"), ~ as.numeric(!is.na(.x)), .names = "missing_{.col}")) %>%
  unite("miss_pattern", starts_with("missing_"), remove = FALSE) %>%
  rowwise() %>% 
  mutate(miss_count = sum(c_across(starts_with("missing_pain_")))) %>%
  group_by(miss_pattern) %>%
  mutate(pattern_freq = n()) %>%
  ungroup() %>%
  rowwise() %>%
  mutate(miss_pattern = gsub("_", ", ", miss_pattern),
         miss_pattern_label = paste0("Pattern: ", miss_pattern, ", (n=", pattern_freq , ")", collapse = "")) %>%
  pivot_longer(starts_with("pain_"), names_to = "time", names_prefix = "pain_", 
               values_to = "pain") %>%
  mutate(
    miss_obs = as.factor(if_else(is.na(pain),0,1)),
    time_cont = case_when(
      time=="base" ~ 0,
      time=="yr1" ~ 1,
      time=="yr2" ~ 2,
      time=="yr3" ~ 3,
      time=="yr4" ~ 4
    )) %>%
  arrange(desc(miss_pattern), Subject_ID) %>%
  mutate(miss_pattern_label = factor(miss_pattern_label, levels = unique(miss_pattern_label)),
         group_vitd = factor(vitdactive, levels = c(0,1), labels = c("Placebo","Vitamin D")),
         group_oil = factor(fishoilactive, levels = c(0,1), labels = c("Placebo","Fish Oil")))
```

```{r mytable, echo=FALSE, fig.cap='Proportions of Missing Data over time in longitudinal trial (MacFarlane et al. 2020)', fig.height=6, fig.width=14}
## Ploting

pct_format = scales::label_percent(accuracy = .1)
suppressMessages(
data_miss %>%
  group_by(time, miss_obs) %>%
  summarise(count_pat = n(), .groups="keep") %>%
  group_by(time) %>%
  mutate(total_pat = sum(count_pat), perc_pat = pct_format(count_pat/total_pat)) %>%
  ungroup %>%
  ggplot(aes(x = time, fill = miss_obs, y = count_pat)) +
  geom_col() +  
  stat_identity(geom = "text", colour = "white", size = 10,
                aes(label =perc_pat), 
                position = position_stack(vjust=0.5)) +
scale_fill_manual(name="",values=c("olivedrab", "lawngreen"), labels=c("Missing Data", "Measured Data")) +
  scale_x_discrete(name = "Time", labels = c("base" = "Baseline", "yr1" = "Year 1", "yr2" = "Year 2", "yr3" = "Year 3", "yr4"="Year 4")) + 
  scale_y_continuous(name = "Number of Patients") +
  theme(text = element_text(size = 18),
        axis.text.x = element_text(angle = 0, vjust = 0.5),
        legend.position = "top", legend.text=element_text(size=20))
)
```


```{r mytable2, echo=FALSE, warning=FALSE, fig.cap='Proportions of missing data(dark green) overtime within treatment groups. Proportions are similar within each group.', fig.height=6, fig.width=14}

## Second plot
suppressMessages(
data_miss %>%
  group_by(time, miss_obs, group_oil) %>%
  summarise(count_pat = n(), .groups="keep") %>%
  group_by(time, group_oil) %>%
  mutate(total_pat = sum(count_pat), perc_pat = pct_format(count_pat/total_pat)) %>%
  ungroup %>%
  ggplot(aes(x = time, fill = miss_obs, y = count_pat)) +
  geom_col() +
  facet_grid(~group_oil) + 
  stat_identity(geom = "text", colour = "white", size = 10,
                aes(label =perc_pat), 
                position = position_stack(vjust=0.5)) +
  scale_fill_manual(name="Data",values=c("olivedrab", "lawngreen"), labels=c("Missing", "Measured"))+
  scale_x_discrete(name = "Time", labels = c("base" = "Baseline", "yr1" = "Year 1", "yr2" = "Year 2", "yr3" = "Year 3", "yr4"="Year 4")) + 
  scale_y_continuous(name = "Number of Patients") +
  theme(text = element_text(size = 18),
        axis.text.x = element_text(angle = 0, vjust = 0.5), legend.position = "none")
)
```
 
```{r mytable3, echo=FALSE, warning=FALSE, message=FALSE, fig.cap='The missing data patterns grouped by fish oil intake. 0/1 indicating data missing/observed. Differences in patterns can be observed.', fig.height=6, fig.width=14}

## Third plot
data_miss_count <- data_miss %>%
  select(Subject_ID, miss_pattern, pain, time_cont, group_vitd, group_oil, miss_obs) %>%
  group_by(group_vitd, group_oil, miss_pattern, miss_obs, time_cont) %>%
  summarise(pattern_count = n()) %>%
  ungroup %>%
  mutate(pattern_perc = round(pattern_count/sum(pattern_count)*100*5,1))

ggplot(data_miss_count, aes(x = time_cont, y = miss_pattern, fill = miss_obs)) +
  geom_tile(color = "white") + 
  geom_text(data = filter(data_miss_count, time_cont == 4), 
            aes(x = time_cont+1.5, label = paste0(pattern_count," (", pattern_perc, "%)")), size = 4) +
  facet_grid(~group_oil, scales = "free", switch = "y") +
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.spacing.x = unit(1, "lines"),
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 90, vjust = 0.5)) +
  coord_cartesian(xlim = c(-0.5, 6.2), clip = 'off') +
  scale_y_discrete(name = "Missing Data Pattern") +
  scale_fill_discrete(name = "Data Observed\nIndicator (R)", ) +
  scale_x_continuous(name = "Time", breaks = c(0,1,2,3,4,5), 
                     labels = c("Baseline","Year 1","Year 2","Year 3","Year 4","Patient Count\n(%)")) +
  scale_fill_manual(name="Treatment Group",values=c("olivedrab", "lawngreen"))
```

# Acupuncture Trial

```{r}
# Formating data

library(readr)
library(ggplot2)
acupuncture <- read_csv("Data/acupuncture.csv",show_col_types = FALSE)
acupuncture_miss <- acupuncture %>%
  select(id, group, pk1, pk2, pk5) %>%
  mutate(across(starts_with("pk"), ~ as.numeric(!is.na(.x)), .names = "missing_{.col}")) %>%
  unite("miss_pattern", starts_with("missing_"), remove = FALSE) %>%
  rowwise() %>% 
  mutate(miss_count = sum(c_across(starts_with("missing_pain_")))) %>%
  group_by(miss_pattern) %>%
  mutate(pattern_freq = n()) %>%
  ungroup() %>%
  rowwise() %>%
  mutate(miss_pattern = gsub("_", ", ", miss_pattern),
         miss_pattern_label = paste0("Pattern: ", miss_pattern, ", (n=", pattern_freq , ")", collapse = "")) %>%
  pivot_longer(starts_with("pk"), names_to = "time", names_prefix = "pk", 
               values_to = "pk") %>%
  mutate(
    miss_obs = as.factor(if_else(is.na(pk),0,1)),
    time_cont = case_when(
      time==1 ~ 0,
      time==2 ~ 1,
      time==5 ~ 2
    )) %>%
  arrange(desc(miss_pattern), id) %>%
  mutate(miss_pattern_label = factor(miss_pattern_label, levels = unique(miss_pattern_label)))
```
```{r mytable4, echo=FALSE, warning=FALSE, fig.cap='Proportions of Missing data over time in longitudinal trial (Vickers et al. 2004)', fig.height=6, fig.width=14}
# 1st plot

pct_format = scales::label_percent(accuracy = .1)

acupuncture_miss %>%
  group_by(time, miss_obs) %>%
  summarise(count_pat = n(), .groups = "keep") %>%
  group_by(time) %>%
  mutate(total_pat = sum(count_pat), perc_pat = pct_format(count_pat/total_pat)) %>%
  ungroup %>%
  ggplot(aes(x = time, fill = miss_obs, y = count_pat)) +
  geom_col() +
  stat_identity(geom = "text", colour = "white", size = 10,
                aes(label =perc_pat), 
                position = position_stack(vjust=0.5)) +
  scale_fill_manual(name="",values=c("olivedrab", "lawngreen"), labels=c("Missing Data", "Measured Data")) + 
  scale_x_discrete(name = "Time", labels = c("1" = "Baseline", "2" = "3 Months", "5" = "12 Months")) +
  scale_y_continuous(name = "Number of Patients") +
  theme(text = element_text(size = 18),
        axis.text.x = element_text(angle = 0, vjust = 0.5), legend.position = "top", legend.text=element_text(size=20))
```

```{r mytable5, echo=FALSE, warning=FALSE, fig.cap='Proportions of missing data(dark green) within treatment groups over time', fig.height=6, fig.width=14}
# 2nd plot

acupuncture_miss %>%
  group_by(time, miss_obs, group) %>%
  summarise(count_pat = n(), .groups="keep") %>%
  group_by(time, group) %>%
  mutate(total_pat = sum(count_pat), perc_pat = pct_format(count_pat/total_pat)) %>%
  ungroup %>%
  ggplot(aes(x = time, fill = miss_obs, y = count_pat)) +
  geom_col() + 
  facet_grid(.~group, labeller=labeller(group=c("0" = "Placebo", "1" = "Acupuncture"))) + 
  stat_identity(geom = "text", colour = "white", size = 10,
                aes(label =perc_pat), 
                position = position_stack(vjust=0.5)) +
  scale_fill_manual(name="Data",values=c("olivedrab", "lawngreen"), labels=c("Missing", "Measured")) + 
 scale_x_discrete(name = "Time", labels = c("1" = "Baseline", "2" = "3 Months", "5" = "12 Months")) +
  scale_y_continuous(name = "Number of Patients") +
  theme(text = element_text(size = 18),
        axis.text.x = element_text(angle = 0, vjust = 0.5),legend.position = "none",strip.text.x=element_text(size=20))
```

```{r mytable6, echo=FALSE, warning=FALSE, fig.cap='Missing data (dark green) pattern over time ', fig.height=6, fig.width=14}
## Third plot

acupuncture_miss_count <- acupuncture_miss %>%
  select(id, miss_pattern, pk, time_cont, group, miss_obs) %>%
  group_by(group, miss_pattern, miss_obs, time_cont) %>%
  summarise(pattern_count = n(), .groups="keep") %>%
  ungroup %>%
  mutate(pattern_perc = round(pattern_count/sum(pattern_count)*100*5,1))


ggplot(acupuncture_miss_count, aes(x = time_cont, y = miss_pattern, fill = miss_obs)) +
  geom_tile(color = "white") + 
  geom_text(data = filter(acupuncture_miss_count, time_cont == 2), 
            aes(x = time_cont+1, label = paste0(pattern_count," (", pattern_perc, "%)")), size = 5.5) +
  facet_grid(~group, scales = "free", switch = "y",labeller=labeller(group=c("0" = "Placebo", "1" = "Acupuncture"))) +
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.spacing.x = unit(1, "lines"),
        text = element_text(size = 18),
        axis.text.x = element_text(angle = 0, vjust = 0.5), legend.position = "none",strip.text.x=element_text(size=20)) +
  coord_cartesian(xlim = c(-0.5, 3.2), clip = 'off') +
  scale_y_discrete(name = "Missing Data Pattern") +
  scale_fill_manual(name = "", labels=c("Missing","Measured"), values=c("olivedrab","lawngreen")) +
  scale_x_continuous(name = "Time", breaks = c(0.1,1.1,2.1,3.1), 
                     labels = c("Baseline","3 Months","12 Months","Patients \n(%)")) 
```

####

# Common Analysis for Data with Missing Observations [^3]
1. Complete case study, participants with missing data are simply excluded.
2. Simple imputation methods(last/baseline observation carried forward)
3. Weigh estimate (not suit in both data set due to non-monotonicity)
4. Multi-imputation & MLE

# Next step

You should mention that, for both datasets, the standard or reference analysis will be via maximum-likelihood, with no imputation. This will involve using linear mixed-effects models to capture the longitudinal aspect. For your work extending this via sensitivity analyses, imputation of one kind or another will be employed in conjunction with a linear mixed-effects model.

Multiple imputation methods will need to consider the longitudinal aspect of the data (observations are correlated within patient). This is a key aspect of the particular design of both trials (as above) and is what makes your project more interesting, beyond a basic multiple imputation approach.


# References
[^1]: MacFarlane, L. A. et al.,(2020). The Effects of Vitamin D and Marine Omega-3 Fatty Acid Supplementation on Chronic Knee Pain in Older US Adults: Results From a Randomized Trial. Arthritis & rheumatology (Hoboken, N.J.), 72(11), 1836–1844. https://doi.org/10.1002/art.41416
[^2]: Vickers, A. J. et al., (2004). Acupuncture for chronic headache in primary care: large, pragmatic, randomised trial. BMJ (Clinical research ed.), 328(7442), 744. https://doi.org/10.1136/bmj.38029.421863.EB
[^3]: Carpenter, J., & Kenward, M. (2007). Missing data in randomised controlled trials: a practical guide. Health Technology Assessment Methodology Programme. 
The code and datasets for this project can be viewed at our GitHub repository here: <https://github.com/amydebrun/Missing-Data.git>
